------------------------------------------------------------------------
title: "PITS_statistical_tests" 
format: html 
editor: visual 
---

### read in libraries

```{r}
library(readr)
library(tidyverse)
library(broom)       # tidy model output
library(MASS)        # glm.nb for negative binomial
library(dunn.test)   # Dunn post-hoc after Kruskal-Wallis
```

### read in data
```{r}
PITS_public <- read_csv("processed_data/PITS_public_share.csv") 
# %>% select(-1)
MyPC_public  <- read_csv("processed_data/MyPC_public_share.csv") 
# %>% select(-1)

```

# =============================================================================
# Clean PITS
# =============================================================================
```{r}
PITS_public <- PITS_public %>%
  mutate(
    # Convert string "NA" to real NA
    `Suspension Status` = na_if(`Suspension Status`, "NA"),
    `Suspension Type`   = na_if(`Suspension Type`,   "NA"),
    Susp_Start_date     = na_if(as.character(Susp_Start_date), "NA"),
    Susp_End_date       = na_if(as.character(Susp_End_date),   "NA"),
    Susp_Start_date     = as.Date(Susp_Start_date),
    Susp_End_date       = as.Date(Susp_End_date),
    Date2               = as.Date(Date2),

    Suspension_length = case_when(
      is.na(Susp_End_date) | is.na(Susp_Start_date) ~ NA_real_,
      Susp_End_date < Susp_Start_date               ~ NA_real_,
      TRUE ~ as.numeric(Susp_End_date - Susp_Start_date)
    ),

    # Drive susp_binary off Suspension Type — avoids "NONE" miscount
    susp_binary = ifelse(`Suspension Type` == "Suspended customer", 1L, 0L),
    susp_binary = replace_na(susp_binary, 0L),

    hour_label = fct_reorder(as.factor(hour_label), hour_label)
  )
PITS_public$PITS_weekday <- weekdays(PITS_public$Date2)

```

MyPC data
```{r}
MyPC_public <- MyPC_public %>%
  mutate(
    booking_date = as.Date(booking_date),
    booking_from = as.Date(booking_from),
    hour_label   = fct_reorder(as.factor(hour_label), hour_label)
  )
MyPC_public$MyPC_weekday <- weekdays(MyPC_public$booking_date)

```


# =============================================================================
# Pre-compute ratio data (computer users vs incidents)
# =============================================================================
```{r}
PITS_incident_counts <- PITS_public %>%
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(incident_count = n_distinct(`Incident ID`), .groups = "drop")

MyPC_counts <- MyPC_public %>%
  group_by(Location_masked, date = booking_date) %>%
  summarise(user_count = n(), .groups = "drop")

ratio_data <- MyPC_counts %>%
  full_join(PITS_incident_counts, by = c("Location_masked", "date")) %>%
  mutate(
    user_count     = replace_na(user_count,     0),
    incident_count = replace_na(incident_count, 0)
  )

```


# =============================================================================
# Pre-compute suspension ratio data (suspensions per incident + avg length)
# =============================================================================
```{r}

# One row per unique suspended incident — prevents duration double-counting
PITS_susp_deduped <- PITS_public %>%
  filter(susp_binary == 1,
         !is.na(`Incident ID`),
         !is.na(Suspension_length)) %>%
  distinct(`Incident ID`, Location_masked, Date2, Suspension_length)

# Suspensions per location per day
PITS_susp_counts <- PITS_susp_deduped %>%
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(
    suspension_count   = n_distinct(`Incident ID`),
    avg_susp_length    = mean(Suspension_length),
    median_susp_length = median(Suspension_length),
    .groups = "drop"
  )

susp_ratio_data <- PITS_incident_counts %>%
  left_join(PITS_susp_counts, by = c("Location_masked", "date")) %>%
  mutate(
    suspension_count   = replace_na(suspension_count,   0),
    avg_susp_length    = replace_na(avg_susp_length,    0),
    median_susp_length = replace_na(median_susp_length, 0),
    susp_ratio         = ifelse(incident_count > 0,
                                suspension_count / incident_count, 0)
  )

```



# =============================================================================
# SECTION 1: Baseline — Suspensions as the dominant outcome
# =============================================================================

## Descriptive: suspension rate overall and by location
```{r}
suspension_summary <- PITS_public %>%
  summarise(
    total_incidents    = n_distinct(`Incident ID`),
    total_suspensions  = sum(susp_binary, na.rm = TRUE),
    suspension_rate    = mean(susp_binary, na.rm = TRUE)
  )

suspension_summary_by_loc <- PITS_public %>%
  group_by(Location_masked) %>%
  summarise(
    total_incidents   = n_distinct(`Incident ID`),
    total_suspensions = sum(susp_binary, na.rm = TRUE),
    suspension_rate   = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(suspension_rate))

print(suspension_summary)
print(suspension_summary_by_loc)
```

## One-sample proportion test: Is suspension rate > 50%?
```{r}
# H0: true suspension rate = 0.50
# H1: true suspension rate > 0.50
binom_test <- binom.test(
  x = suspension_summary$total_suspensions,
  n = suspension_summary$total_incidents,
  p = 0.50,
  alternative = "greater"
)
print(binom_test)
```

## Support service use vs. suspension rate
```{r}
#| warning: false
#| message: false

support_summary <- PITS_public %>%
  summarise(
    n                  = n(),
    pct_suspended      = mean(susp_binary, na.rm = TRUE),
    pct_guard_called   = mean(`Support:Called Guard?`         == "Yes", na.rm = TRUE),
    pct_police_called  = mean(`Support:Called Police (911)?`  == "Yes", na.rm = TRUE),
    pct_emt_called     = mean(`Support:Called EMT?`           == "Yes", na.rm = TRUE),
    pct_crisis_called  = mean(`Support:Called crisis line?`   == "Yes", na.rm = TRUE)
  ) %>%
  pivot_longer(-n, names_to = "response_type", values_to = "rate") %>%
  mutate(response_type = gsub("pct_", "", response_type))

print(support_summary)

# McNemar test: suspension vs. guard called (paired binary comparison)
# Requires a contingency table of the two binary variables
mcnemar_df <- PITS_public %>%
  mutate(
    guard_called = ifelse(`Support:Called Guard?` == "Yes", 1L, 0L),
    guard_called = replace_na(guard_called, 0L)
  )

mcnemar_table <- table(
  suspended    = mcnemar_df$susp_binary,
  guard_called = mcnemar_df$guard_called
)

print(mcnemar_table)
mcnemar_result <- mcnemar.test(mcnemar_table)
print(mcnemar_result)
```


# =============================================================================
# SECTION 2 (Main): Logistic regression — does reporting predict suspension
#                   better than incident characteristics do?
# =============================================================================

```{r, echo=FALSE}
#| warning: false
#| message: false

# Prepare: drop rows missing key predictors
pits_model_df <- PITS_public %>%
  filter(!is.na(susp_binary),
         !is.na(parsed_categories),
         !is.na(hour_label),
         !is.na(Location_masked),
         !is.na(PITS_weekday)) %>%
  mutate(
    infraction_cat = as.factor(parsed_categories),
    location       = as.factor(Location_masked),
    weekday        = as.factor(PITS_weekday),
    hour           = as.factor(hour_label)
  )

# Null model (intercept only — "reporting alone")
logit_null <- glm(susp_binary ~ 1,
                  data   = pits_model_df,
                  family = binomial)

# Full model with incident characteristics
logit_full <- glm(susp_binary ~ infraction_cat + hour + weekday + location,
                  data   = pits_model_df,
                  family = binomial)

summary(logit_full)

# McFadden pseudo-R² — low value supports argument that
# observable characteristics don't explain suspension well
mcfadden_r2 <- 1 - (as.numeric(logLik(logit_full)) /
                      as.numeric(logLik(logit_null)))
cat("McFadden pseudo-R²:", round(mcfadden_r2, 4), "\n")

# Tidy coefficient table
logit_tidy <- tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
  arrange(p.value)
print(logit_tidy)

# Likelihood ratio test: does the full model fit better than null?
lrt_result <- anova(logit_null, logit_full, test = "LRT")
print(lrt_result)
```


# =============================================================================
# SECTION 3: Chi-square — suspensions not proportional to infraction type
# =============================================================================
A non-significant result here supports you: suspensions are equally likely regardless of what the infraction was
```{r}
#| warning: false
#| message: false

# Contingency table: infraction category × suspension binary
infraction_susp_table <- table(
  infraction = PITS_public$parsed_categories,
  suspended  = PITS_public$susp_binary
)

print(infraction_susp_table)

chi_infraction <- chisq.test(infraction_susp_table)
print(chi_infraction)

# Expected vs observed suspension rate per infraction category
infraction_rates <- PITS_public %>%
  filter(!is.na(parsed_categories)) %>%
  group_by(parsed_categories) %>%
  summarise(
    n               = n(),
    n_suspended     = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(suspension_rate))

print(infraction_rates)
```


# =============================================================================
# SECTION 4: Kruskal-Wallis — suspension length not calibrated to infraction
# =============================================================================
```{r}
#| warning: false
#| message: false

# Proportion of incidents by infraction category
PITS_public %>%
  filter(!is.na(parsed_categories)) %>%
  count(parsed_categories) %>%
  mutate(pct = n / sum(n) * 100) %>%
  arrange(desc(n))

# Suspension rate *within* each category
# (are large categories also suspending at higher rates?)
PITS_public %>%
  filter(!is.na(parsed_categories)) %>%
  group_by(parsed_categories) %>%
  summarise(
    n               = n(),
    n_suspended     = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(n))


```

Tests whether suspension length varies by infraction category — if it doesn't, lengths aren't being calibrated to severity either
```{r}
# Filter to suspended incidents only, with valid length
susp_length_df <- PITS_public %>%
  filter(susp_binary == 1,
         !is.na(Suspension_length),
         !is.na(parsed_categories))

# Descriptive: median length by infraction category
length_by_infraction <- susp_length_df %>%
  group_by(parsed_categories) %>%
  summarise(
    n              = n(),
    median_days    = median(Suspension_length),
    mean_days      = mean(Suspension_length),
    sd_days        = sd(Suspension_length),
    .groups = "drop"
  ) %>%
  arrange(desc(median_days))

print(length_by_infraction)

# Kruskal-Wallis test
kw_length <- kruskal.test(
  Suspension_length ~ as.factor(parsed_categories),
  data = susp_length_df
)
print(kw_length)

# Dunn post-hoc test (only meaningful if KW is significant,
# but run regardless — non-significant pairs support your argument)
dunn_length <- dunn.test(
  x     = susp_length_df$Suspension_length,
  g     = susp_length_df$parsed_categories,
  method = "bonferroni",
  kw    = TRUE
)

# Kruskal-Wallis by location
kw_location <- kruskal.test(
  Suspension_length ~ as.factor(Location_masked),
  data = susp_length_df
)
print(kw_location)
```


# =============================================================================
# SECTION 5 (Supporting): Poisson/NB regression — usage volume does not
#                          predict incident counts
# =============================================================================

```{r}
#| warning: false

# Poisson model: do more computer users → more incidents?
poisson_model <- glm(
  incident_count ~ user_count + Location_masked,
  data   = ratio_data,
  family = poisson
)
summary(poisson_model)

# Check for overdispersion (variance >> mean → use negative binomial)
dispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) /
                    poisson_model$df.residual
cat("Dispersion ratio:", round(dispersion_ratio, 3),
    " (>1.5 suggests overdispersion → prefer NB)\n")

# Negative binomial model (more robust if overdispersed)
nb_model <- glm.nb(
  incident_count ~ user_count + Location_masked,
  data = ratio_data
)
summary(nb_model)

# Tidy both for comparison
tidy_poisson <- tidy(poisson_model, exponentiate = TRUE, conf.int = TRUE)
tidy_nb      <- tidy(nb_model,      exponentiate = TRUE, conf.int = TRUE)

cat("\n--- Poisson IRRs ---\n"); print(tidy_poisson)
cat("\n--- Negative Binomial IRRs ---\n"); print(tidy_nb)
```


# =============================================================================
# SECTION 6: Chi-square — suspension rates vary by location (inconsistent
#            application)
# =============================================================================

```{r}
#| warning: false
#| message: false

# Contingency table: location × suspension binary
location_susp_table <- table(
  location  = PITS_public$Location_masked,
  suspended = PITS_public$susp_binary
)

print(location_susp_table)

chi_location <- chisq.test(location_susp_table)
print(chi_location)

# Standardised residuals — which locations suspend more/less than expected?
std_residuals <- as.data.frame(chi_location$stdres) %>%
  setNames(c("location", "suspended", "std_resid")) %>%
  filter(suspended == "1") %>%
  arrange(desc(std_resid))

cat("\nStandardised residuals (suspended == 1):\n")
print(std_residuals)

# Logistic regression: location as predictor of suspension
logit_location <- glm(
  susp_binary ~ Location_masked,
  data   = PITS_public,
  family = binomial
)
summary(logit_location)
tidy(logit_location, exponentiate = TRUE, conf.int = TRUE)
```

# =============================================================================
# SECTION 7: Day of week and time of day — incidents and suspensions
# =============================================================================

```{r}
#| warning: false
#| message: false

# ── 7a. Order weekdays Mon–Sun ────────────────────────────────────────────────
weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday",
                    "Friday", "Saturday", "Sunday")

PITS_public <- PITS_public %>%
  mutate(PITS_weekday = factor(PITS_weekday, levels = weekday_levels))

# # ── 7b. Incidents and suspensions by day of week ──────────────────────────────
# dow_summary <- PITS_public %>%
#   filter(!is.na(PITS_weekday)) %>%
#   group_by(PITS_weekday) %>%
#   summarise(
#     n_incidents   = n_distinct(`Incident ID`),
#     n_suspensions = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# print(dow_summary)
# 
# # Chi-square: are incidents distributed evenly across days?
# chisq_dow_incidents <- chisq.test(dow_summary$n_incidents)
# cat("\nChi-square — incident counts across weekdays:\n")
# print(chisq_dow_incidents)
# 
# # Chi-square: are suspensions distributed evenly across days?
# chisq_dow_suspensions <- chisq.test(dow_summary$n_suspensions)
# cat("\nChi-square — suspension counts across weekdays:\n")
# print(chisq_dow_suspensions)
# 
# # ── 7c. Suspension rate by day — is any day more likely to suspend? ───────────
# # Logistic regression with weekday as predictor
# logit_dow <- glm(
#   susp_binary ~ PITS_weekday,
#   data   = PITS_public %>% filter(!is.na(PITS_weekday)),
#   family = binomial
# )
# cat("\nLogistic regression — weekday as predictor of suspension:\n")
# tidy(logit_dow, exponentiate = TRUE, conf.int = TRUE) %>% print()

# In Section 7b — filter Sunday out
dow_summary <- PITS_public %>%
  filter(!is.na(PITS_weekday), PITS_weekday != "Sunday") %>%
  group_by(PITS_weekday) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

# In Section 7c — filter Sunday out of the logistic regression too
logit_dow <- glm(
  susp_binary ~ PITS_weekday,
  data   = PITS_public %>% filter(!is.na(PITS_weekday), PITS_weekday != "Sunday"),
  family = binomial
)
# ── 7d. Incidents and suspensions by hour of day ──────────────────────────────
# hour_summary <- PITS_public %>%
#   filter(!is.na(hour_label)) %>%
#   group_by(hour_label) %>%
#   summarise(
#     n_incidents     = n_distinct(`Incident ID`),
#     n_suspensions   = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   arrange(hour_label)



```

```{r}
library(lubridate)


hour_summary <- PITS_public %>%
  filter(!is.na(hour_label)) %>%
  mutate(hour_label = substr(hour_label, 1, 2)) %>%
  group_by(hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(hour_label))


# Kruskal-Wallis: does suspension rate vary by hour?
# (use raw binary per row, not aggregated rates)
kw_hour <- kruskal.test(
  susp_binary ~ as.factor(hour_label),
  data = PITS_public %>% filter(!is.na(hour_label))
)
cat("\nKruskal-Wallis — suspension rate by hour of day:\n")
print(kw_hour)

# ── 7e. Cross-tab: day of week × hour of day (incident counts) ───────────────
dow_hour_table <- PITS_public %>%
  filter(!is.na(PITS_weekday), !is.na(hour_label)) %>%
  group_by(PITS_weekday, hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

# Peak combinations — top 10 by incident count
cat("\nTop 10 day × hour combinations by incident count:\n")
dow_hour_table %>%
  arrange(desc(n_incidents)) %>%
  slice_head(n = 10) %>%
  print()

# Peak combinations — top 10 by suspension rate (min 5 incidents to filter noise)
cat("\nTop 10 day × hour combinations by suspension rate (n ≥ 5):\n")
dow_hour_table %>%
  filter(n_incidents >= 5) %>%
  arrange(desc(suspension_rate)) %>%
  slice_head(n = 10) %>%
  print()
```

Day of Week
Incident counts are not evenly distributed (χ² p < .0001). Tuesday, Thursday, and Wednesday are the busiest days (586, 616, 526 incidents), while Sunday is nearly absent (9 incidents). This likely reflects library operating hours and foot traffic patterns rather than anything behavioral.
The more interesting finding is the suspension rate pattern. Suspension rates fall as the week progresses:
DaySuspension RateMonday48.8%Tuesday43.6%Wednesday43.8%Thursday42.3%Friday35.0%Saturday31.1%Sunday29.4%
The logistic regression confirms this is statistically significant for Tuesday through Saturday (all p < .05), with Monday as the reference category. The odds ratios tell you the direction clearly — by Friday, an incident is only 0.57 times as likely to result in suspension as the same incident on a Monday, and Saturday drops to 0.48. Sunday is not significant, likely because n=9 is too small to detect an effect.
How to frame this: Early-week incidents are substantially more likely to result in suspension than late-week incidents, even though the nature of the infraction hasn't changed. This is consistent with a discretionary application argument — staff behavior around suspension varies systematically by day in a way the written policy doesn't explain or prescribe.

Hour of Day
The early-morning hours are an immediate data quality flag. Hours 00:00–06:00 have very small n (1–10 incidents) but produce extreme suspension rates — 92.9% at 04:00, 78.6% at 02:00. These are almost certainly data entry errors or timestamp anomalies rather than real overnight incidents, since most public libraries are not open at 2–4am. I'd recommend either filtering these out of the analysis or noting them as artifacts.
The meaningful hours are 08:00 onward. The Kruskal-Wallis is significant (χ² = 249, p < .0001), so suspension rates genuinely vary by hour. You'll want to print the full table to see the afternoon pattern clearly.

Day × Hour combinations
Two distinct patterns emerge:
Peak incident volume clusters in weekday afternoons, 13:00–16:00 — Thursday 15:00 (n=92), Wednesday 14:00 (n=88), Thursday 16:00 (n=84). This is your busy period, likely corresponding to after-school hours and afternoon foot traffic.
Peak suspension rates cluster around 17:00 on weekdays — Monday 17:00 (85.4%), Tuesday 17:00 (71.7%), Thursday 17:00 (69.1%). This is striking because 17:00 is not the busiest hour for incidents, but it produces the highest suspension rates. One plausible explanation: 17:00 is approaching closing time at many branches, and staff may be more likely to suspend as a way of resolving a situation before end of shift rather than leaving it open.
Wednesday 16:00 is an outlier worth flagging — 72 incidents but 102 suspensions (suspension rate 60.7%). The fact that suspensions exceed incidents suggests some incidents generated multiple suspension records, which may be worth investigating in your data.

```{r}
incident_hour_table <- PITS_public %>%
  filter(!is.na(PITS_weekday), !is.na(hour_label)) %>%
  group_by(PITS_weekday, hour_label) %>%
  summarise(
    n_incidents = n_distinct(`Incident ID`),
    .groups = "drop"
  ) %>%
  mutate(
    incident_rate = n_incidents / sum(n_incidents)
  ) 

incident_hour_table%>%
  arrange(desc(incident_rate)) %>%
  slice_head(n = 10)
```


```{r}
```


```{r}
incident_hour_table
```

# ── 7f. Visualisations ────────────────────────────────────────────────────────

```{r fig.width=9, fig.height=4}
# Incidents by day of week
ggplot(dow_summary, aes(x = PITS_weekday, y = n_incidents)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n_incidents), vjust = -0.4, size = 3.5) +
  labs(title = "Incidents by Day of Week",
       x = NULL, y = "Number of Incidents") +
  theme_minimal()

# Suspension rate by day of week
ggplot(dow_summary, aes(x = PITS_weekday, y = suspension_rate)) +
  geom_col(fill = "coral") +
  geom_text(aes(label = scales::percent(suspension_rate, accuracy = 1)),
            vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Day of Week",
       x = NULL, y = "Suspension Rate") +
  theme_minimal()
```

```{r fig.width=11, fig.height=4}
# Incidents by hour of day
ggplot(hour_summary, aes(x = hour_label, y = n_incidents)) +
  geom_col(fill = "steelblue") +
  labs(title = "Incidents by Hour of Day",
       x = "Hour", y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Suspension rate by hour of day
ggplot(hour_summary, aes(x = hour_label, y = suspension_rate)) +
  geom_col(fill = "coral") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Hour of Day",
       x = "Hour", y = "Suspension Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=11, fig.height=6}
#| warning: false
#| message: false

# Heatmap: incidents by day × hour
ggplot(dow_hour_table,
       aes(x = hour_label, y = PITS_weekday, fill = n_incidents)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = "steelblue",
                      name = "Incidents") +
  labs(title = "Incident Count Heatmap: Day of Week × Hour of Day",
       x = "Hour", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Heatmap: suspension rate by day × hour (cells with n < 5 greyed out)
dow_hour_table %>%
  mutate(rate_display = ifelse(n_incidents < 5, NA, suspension_rate)) %>%
  ggplot(aes(x = hour_label, y = PITS_weekday, fill = rate_display)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = "coral",
                      name = "Susp. Rate",
                      labels = scales::percent_format(),
                      na.value = "grey90") +
  labs(title = "Suspension Rate Heatmap: Day of Week × Hour of Day",
       subtitle = "Grey = fewer than 5 incidents",
       x = "Hour", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# =============================================================================
# SECTION 8: MyPC usage volume by day of week — are busy days incident days?
# =============================================================================

```{r}
#| warning: false
#| message: false

# ── 8a. Summarise MyPC usage by day of week ───────────────────────────────────
# MyPC_public <- MyPC_public %>%
#   mutate(MyPC_weekday = factor(MyPC_weekday, levels = weekday_levels))
# 
# mypc_dow <- MyPC_public %>%
#   filter(!is.na(MyPC_weekday)) %>%
#   group_by(MyPC_weekday) %>%
#   summarise(
#     n_sessions    = n(),
#     n_unique_users = n_distinct(new_login_id),
#     .groups = "drop"
#   )

# In Section 8a — filter Sunday out of MyPC
mypc_dow <- MyPC_public %>%
  filter(!is.na(MyPC_weekday), MyPC_weekday != "Sunday") %>%
  group_by(MyPC_weekday) %>%
  summarise(
    n_sessions     = n(),
    n_unique_users = n_distinct(new_login_id),
    .groups = "drop"
  )
print(mypc_dow)

# ── 8b. Join MyPC usage with incident counts by weekday ───────────────────────
dow_combined <- dow_summary %>%
  left_join(mypc_dow, by = c("PITS_weekday" = "MyPC_weekday")) %>%
  mutate(
    incidents_per_100_sessions = (n_incidents / n_sessions) * 100,
    suspensions_per_100_sessions = (n_suspensions / n_sessions) * 100
  )

print(dow_combined)

# ── 8c. Correlation: do busier days produce more incidents? ───────────────────
cor_sessions_incidents <- cor.test(
  dow_combined$n_sessions,
  dow_combined$n_incidents,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. incidents by weekday:\n")
print(cor_sessions_incidents)

cor_sessions_suspensions <- cor.test(
  dow_combined$n_sessions,
  dow_combined$n_suspensions,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. suspensions by weekday:\n")
print(cor_sessions_suspensions)

# ── 8d. Correlation: does higher usage predict higher suspension RATE? ─────────
# This is the key test — if suspension rate rises with usage,
# it's a volume effect; if not, something else is driving it
cor_sessions_rate <- cor.test(
  dow_combined$n_sessions,
  dow_combined$suspension_rate,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. suspension rate by weekday:\n")
print(cor_sessions_rate)
```

```{r fig.width = 9, fig.height = 4}
#| warning: false
#| message: false

# ── 8e. Visualise: sessions vs incidents by weekday ───────────────────────────
dow_combined %>%
  dplyr::select(PITS_weekday, n_incidents, n_sessions) %>%
  pivot_longer(cols = c(n_incidents, n_sessions),
               names_to = "metric", values_to = "count") %>%
  mutate(metric = recode(metric,
                         n_incidents = "Incidents",
                         n_sessions  = "Computer Sessions")) %>%
  ggplot(aes(x = PITS_weekday, y = count, fill = metric)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Computer Sessions" = "steelblue",
                                "Incidents"         = "coral")) +
  labs(title = "Computer Sessions vs. Incidents by Day of Week",
       x = NULL, y = "Count", fill = NULL) +
  theme_minimal()

# ── 8f. Incidents per 100 sessions — normalised incident rate ─────────────────
ggplot(dow_combined,
       aes(x = PITS_weekday, y = incidents_per_100_sessions)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(incidents_per_100_sessions, 2)),
            vjust = -0.4, size = 3.5) +
  labs(title = "Incidents per 100 Computer Sessions by Day of Week",
       subtitle = "Normalises for usage volume — higher = more incidents relative to visitors",
       x = NULL, y = "Incidents per 100 Sessions") +
  theme_minimal()

# ── 8g. Suspension rate vs. session volume scatterplot ────────────────────────
ggplot(dow_combined,
       aes(x = n_sessions, y = suspension_rate, label = PITS_weekday)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_text(vjust = -0.7, size = 3.5) +
  geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate vs. Computer Session Volume by Weekday",
       subtitle = "If volume drove suspension rates, points would trend upward",
       x = "Number of Computer Sessions", y = "Suspension Rate") +
  theme_minimal()
```

```{r}
#| warning: false
#| message: false

library(ggrepel)

ggplot(dow_combined,
       aes(x = n_sessions, y = suspension_rate, label = PITS_weekday)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_text_repel(size = 3.5,
                  nudge_y = -0.01,
                  box.padding = 0.4,
                  point.padding = 0.3) +
  geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate vs. Computer Session Volume by Weekday",
       # subtitle = "If volume drove suspension rates, points would trend upward",
       x = "Number of Computer Sessions", y = "Suspension Rate") +
  theme_minimal()
```

1. Add cramer's v test 
```{r}
library(rcompanion)

location_susp_table <- table(
  location  = PITS_public$Location_masked,
  suspended = PITS_public$susp_binary
)

# print(location_susp_table)
# Location × suspension
cramer_location <- cramerV(location_susp_table)
cat("Cramér's V — Location:", cramer_location, "\n")

# Weekday × suspension
weekday_susp_table <- table(
  weekday   = PITS_public$PITS_weekday,
  suspended = PITS_public$susp_binary
)
cramer_weekday <- cramerV(weekday_susp_table)
cat("Cramér's V — Weekday:", cramer_weekday, "\n")

# Hour × suspension
hour_susp_table <- table(
  hour      = PITS_public$hour_label,
  suspended = PITS_public$susp_binary
)
cramer_hour <- cramerV(hour_susp_table)
cat("Cramér's V — Hour:", cramer_hour, "\n")
```


2. Combined logistic regression — compare contribution of each predictor
This is the key test. Put location, weekday, and hour into the same model and see which one contributes the most explanatory power when the others are controlled for.
```{r}
combined_df <- PITS_public %>%
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label))

# Full model
logit_combined <- glm(
  susp_binary ~ Location_masked + PITS_weekday + hour_label,
  data   = combined_df,
  family = binomial
)
summary(logit_combined)

# Drop each block and compare fit via likelihood ratio test
logit_no_location <- glm(susp_binary ~ PITS_weekday + hour_label,
                         data = combined_df, family = binomial)
logit_no_weekday   <- glm(susp_binary ~ Location_masked + hour_label,
                         data = combined_df, family = binomial)
logit_no_hour       <- glm(susp_binary ~ Location_masked + PITS_weekday,
                         data = combined_df, family = binomial)

cat("\nDoes removing Location hurt the model? (larger stat = location matters more)\n")
anova(logit_no_location, logit_combined, test = "LRT")

cat("\nDoes removing Weekday hurt the model?\n")
anova(logit_no_weekday, logit_combined, test = "LRT")

cat("\nDoes removing Hour hurt the model?\n")
anova(logit_no_hour, logit_combined, test = "LRT")
```
collapse spase hours
```{r}
combined_df %>%
  count(Location_masked, susp_binary) %>%
  pivot_wider(names_from = susp_binary, values_from = n, values_fill = 0) %>%
  filter(`1` < 3 | `0` < 3)  # flag branches with near-zero cells

combined_df %>%
  count(hour_label, susp_binary) %>%
  pivot_wider(names_from = susp_binary, values_from = n, values_fill = 0) %>%
  filter(`1` < 3 | `0` < 3)  # flag hours with near-zero cells
```

Next: Cramér's V and pseudo-R² for the final piece
Once the separation issue is addressed, re-run:
```{r}
mcfadden <- function(model, null) {
  1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
}

logit_null     <- glm(susp_binary ~ 1, data = combined_df, family = binomial)
logit_location <- glm(susp_binary ~ Location_masked, data = combined_df, family = binomial)
logit_weekday  <- glm(susp_binary ~ PITS_weekday,     data = combined_df, family = binomial)
logit_hour     <- glm(susp_binary ~ hour_label,        data = combined_df, family = binomial)

cat("Pseudo-R² — Location alone:", round(mcfadden(logit_location, logit_null), 4), "\n")
cat("Pseudo-R² — Weekday alone: ", round(mcfadden(logit_weekday, logit_null), 4), "\n")
cat("Pseudo-R² — Hour alone:    ", round(mcfadden(logit_hour, logit_null), 4), "\n")
```

3. compare variance explained with pseudo R^2 by predictor alone
```{r}
logit_null      <- glm(susp_binary ~ 1, data = combined_df, family = binomial)
logit_location  <- glm(susp_binary ~ Location_masked, data = combined_df, family = binomial)
logit_weekday   <- glm(susp_binary ~ PITS_weekday,     data = combined_df, family = binomial)
logit_hour      <- glm(susp_binary ~ hour_label,        data = combined_df, family = binomial)

mcfadden <- function(model, null) {
  1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
}

cat("Pseudo-R² — Location alone:", round(mcfadden(logit_location, logit_null), 4), "\n")
cat("Pseudo-R² — Weekday alone: ", round(mcfadden(logit_weekday, logit_null), 4), "\n")
cat("Pseudo-R² — Hour alone:    ", round(mcfadden(logit_hour, logit_null), 4), "\n")
```
4.  Mixed-effects model — treat location as a random effect
If you want to go further, a mixed model quantifies how much of the unexplained variance in suspension is attributable to branch-level clustering, independent of time effects.
```{r}
library(lme4)

mixed_model <- glmer(
  susp_binary ~ PITS_weekday + hour_label + (1 | Location_masked),
  data   = combined_df,
  family = binomial
)
summary(mixed_model)

# Intraclass correlation — how much variance is due to branch?
library(performance)
icc(mixed_model)
```
deviations from the average log-odds of suspension
```{r}
library(lme4)

ranef_branch <- ranef(mixed_model)$Location_masked
ranef_branch$Location_masked <- rownames(ranef_branch)
names(ranef_branch)[1] <- "random_effect"

ranef_branch <- ranef_branch %>%
  arrange(desc(random_effect))

print(ranef_branch)

# Visualize
ggplot(ranef_branch, aes(x = reorder(Location_masked, random_effect),
                         y = random_effect)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Branch-Level Random Effects (Suspension Likelihood)",
       subtitle = "Positive = higher than average odds of suspension, controlling for time",
       x = NULL, y = "Random Effect (log-odds deviation)") +
  theme_minimal()
```

