---
title: "PITS_suspensions_analysis"
format: html
editor: visual
---
### read in libraries
```{r}
library(readr)
library(tidyverse)
```

### read in data
```{r}
PITS_public <- read_csv("processed_data/PITS_public2.csv") %>% select(-1)
PITS_public <- PITS_public %>% 
  mutate(hour_label = fct_reorder(as.factor(hour_label), hour_label)) 

MyPC_public <- read_csv("processed_data/MyPC_public.csv") %>% select(-1) 
MyPC_public <- MyPC_public %>% 
  mutate(hour_label = fct_reorder(as.factor(hour_label), hour_label)) 
```

hourly grouped usage
```{r}
hourly_usage <- MyPC_public %>%
  group_by(Location_Standard, hour_label, booking_date) %>%
  summarize(count = n(), .groups = "drop") %>% 
  arrange(as.numeric(substr(hour_label, 1, 2)))
# mutate(hour_label = fct_reorder(as.factor(hour_label), hour_label)) 

levels(hourly_usage$hour_label)
```
# streamlined ratio process #
```{r}
# for date, computer usage and any infractions
PITS_counts <- PITS_public %>%
  mutate(date = as.Date(Date2)) %>%
  group_by(Location_masked, date) %>%
  summarize(infraction_count = n(), .groups = "drop")

# Aggregate MyPC data by location and date
MyPC_counts <- MyPC_public %>%
  rename(date = booking_date) %>%
  group_by(Location_masked, date) %>%
  summarize(user_count = n(), .groups = "drop")

# Step 2: Join the datasets
ratio_data <- MyPC_counts %>%
  full_join(PITS_counts, by = c("Location_masked", "date")) %>%
  mutate(
    infraction_count = ifelse(is.na(infraction_count), 0, infraction_count),
    user_count = ifelse(is.na(user_count), 0, user_count),
    infraction_pct = round(
  ifelse(user_count > 0, infraction_count / user_count * 100, NA_real_), 2)
  ) 

```

# for suspensions
```{r}
# for date, computer usage and any infractions
# Total incidents per location per day
PITS_incident_counts <- PITS_public %>%
  group_by(`Incident ID`, Location_masked, Date2) %>%
  summarize(incident_count = n_distinct(`Incident ID`), .groups = "drop")

# Total suspensions per location per day
PITS_susp_counts <- PITS_public %>%
  filter(susp_binary == 1) %>%
  group_by(Location_masked, Date2) %>%
  summarize(suspension_count = n_distinct(`Incident ID`), .groups = "drop")

# Join and compute ratio — both already at Location_masked + Date2 grain
susp_ratio_data <- PITS_incident_counts %>%
  left_join(PITS_susp_counts, by = c("Location_masked", "Date2")) %>%
  mutate(
    suspension_count = ifelse(is.na(suspension_count), 0, suspension_count),
    susp_ratio       = ifelse(incident_count > 0, 
                              suspension_count / incident_count, 
                              0)
  #   suspension_pct = round(
  # ifelse(incident_count > 0, suspension_count / incident_count * 100, NA_real_), 2)
  )

```

looking at suspension length
```{r}
# One row per suspended incident — deduplicate before averaging
PITS_susp_deduped <- PITS_public %>%
  filter(susp_binary == 1, !is.na(Suspension_length), !is.na(`Incident ID`)) %>%
  distinct(`Incident ID`, Location_masked, Date2, Suspension_length)

# Average suspension length per location per day
PITS_susp_avg <- PITS_susp_deduped %>%
  group_by(Location_masked, Date2) %>%
  summarize(
    suspension_count  = n_distinct(`Incident ID`),
    avg_susp_length   = mean(Suspension_length),
    # median_susp_length = median(Suspension_length),
    .groups = "drop"
  )

# Join back to incident counts for full picture
susp_ratio_data <- PITS_incident_counts %>%        # from your previous step
  left_join(PITS_susp_avg, by = c("Location_masked", "Date2")) %>%
  mutate(
    suspension_count   = ifelse(is.na(suspension_count), 0, suspension_count),
    avg_susp_length    = ifelse(is.na(avg_susp_length), 0, avg_susp_length),
    # median_susp_length = ifelse(is.na(median_susp_length), 0, median_susp_length),
    susp_ratio         = ifelse(incident_count > 0,
                                suspension_count / incident_count, 0)
  )
```

```{r}
PITS_incident_counts <- PITS_public %>%
  group_by(Location_masked, Date2) %>%
  summarize(incident_count = n_distinct(`Incident ID`), .groups = "drop")

susp_ratio_data <- PITS_incident_counts %>%
  left_join(PITS_susp_counts, by = c("Location_masked", "Date2")) %>%
  mutate(
    suspension_count = ifelse(is.na(suspension_count), 0, suspension_count),
    susp_ratio       = round(ifelse(incident_count > 0, 
                                    suspension_count / incident_count, NA_real_), 4),
    susp_pct         = round(susp_ratio * 100, 4)
  )

```

Graph 
```{r}
library(lubridate)
library(ggplot2)

monthly_susp <- PITS_public %>%
  mutate(month = floor_date(as.Date(Date2), "month")) %>%
  filter(month >= as.Date("2023-01-01"), month <= as.Date("2025-12-31")) %>%
  group_by(month) %>%
  summarize(
    incident_count   = n_distinct(`Incident ID`),
    suspension_count = sum(susp_binary == 1, na.rm = TRUE),
    susp_pct         = round(suspension_count / incident_count * 100, 4),
    .groups = "drop"
  )


# View the table
print(monthly_susp)

# Plot
ggplot(monthly_susp, aes(x = month, y = susp_pct)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
  scale_y_continuous(labels = scales::percent_format(scale = 1, accuracy = 0.01)) +
  labs(
    title = "Monthly % of Infractions Resulting in Suspension",
    x = NULL, y = "Suspension Rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# ggplot(monthly_susp, aes(x = month, y = susp_pct)) +
#   geom_line(color = "steelblue", linewidth = 1) +
#   geom_point(color = "steelblue", size = 2) +
#   scale_x_date(date_breaks = "3 months", date_labels = "%b %Y") +
#   scale_y_continuous(labels = scales::percent_format(scale = 1, accuracy = 0.01)) +
#   coord_cartesian(ylim = c(0, 100)) +  # Add this
#   labs(
#     title = "Monthly % of Infractions Resulting in Suspension",
#     x = NULL, y = "Suspension Rate (%)"
#   ) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
yearly_susp <- PITS_public %>%
  mutate(year = floor_date(as.Date(Date2), "year")) %>%
  filter(year >= as.Date("2023-01-01"), year <= as.Date("2025-12-31")) %>%
  group_by(year) %>%
  summarize(
    incident_count   = n_distinct(`Incident ID`),
    suspension_count = sum(susp_binary == 1, na.rm = TRUE),
    susp_pct         = round(suspension_count / incident_count * 100, 4),
    .groups = "drop"
  )
yearly_susp
```
save reproducible copies of the datafiles
```{r}
names(MyPC_public)
```
drop site, location, resource, location_standard, location
create a new login_id
```{r}
# Install and load the digest package
if(!require(digest)) install.packages("digest")
library(digest)

# # Sample Data
# df <- data.frame(
#   original_id = c("user123", "user456", "user123", "user789"),
#   data = c(10, 20, 15, 30)
# )

# Generate new_login_id using MD5 hash
MyPC_public$new_login_id <- sapply(MyPC_public$`Login ID`, digest, algo="md5")

```
# Drop the original ID column
# df$original_id <- NULL
drop site, location, resource, location_standard, location, login id
```{r}
MyPC_public_share <- MyPC_public %>% 
  select(-c(Location, Location_Standard, `Login ID`, Site, Resource))

```

```{r}
names(PITS_public)
```

```{r}
PITS_public_share <- PITS_public %>% 
  select(-Location_Standard)
```

write csv
```{r}
write.csv(PITS_public_share, "processed_data/PITS_public_share.csv")

write.csv(MyPC_public_share, "processed_data/MyPC_public_share.csv")
```

