---
title: "PITS_visual_output"
format: html
editor: visual
---
```{r}
library(ggplot2)
library(tidyverse)
```


# ── 7f. Visualisations ────────────────────────────────────────────────────────

```{r fig.width=9, fig.height=4}
# Incidents by day of week
dow_summary <- PITS_deduplicated %>%
  filter(!is.na(PITS_weekday)) %>%
  group_by(PITS_weekday) %>%
  summarise(
    n_incidents   = n_distinct(`Incident ID`),
    n_suspensions = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )
dow_summary_exc <- dow_summary %>% 
      filter(!is.na(PITS_weekday), PITS_weekday != "Sunday") %>%
      mutate(PITS_weekday = factor(PITS_weekday, 
                          levels = c("Monday", "Tuesday","Wednesday",
                                     "Thursday", "Friday","Saturday"), 
                          ordered = TRUE))

ggplot(dow_summary_exc, aes(x = PITS_weekday, y = n_incidents)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = n_incidents), vjust = -0.4, size = 3.5) +
  labs(title = "Incidents by Day of Week",
       x = NULL, y = "Number of Incidents") +
  theme_minimal() # dow_summary is based off of deduped data

# Suspension rate by day of week
ggplot(dow_summary_exc, aes(x = PITS_weekday, y = suspension_rate)) +
  geom_col(fill = "coral") +
   # geom_text(aes(label = n_suspensions), vjust = -0.4, size = 3.5) +
  geom_text(aes(label = scales::percent(suspension_rate, accuracy = 1)),
            vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Day of Week",
       x = NULL, y = "Suspension Rate") +
  theme_minimal()
```



```{r fig.width=11, fig.height=4}
# Incidents by hour of day
hour_summary <- PITS_deduplicated %>%
  filter(!is.na(hour_label)) %>%
  mutate(
    hour_display = format(strptime(hour_label, format = "%H:%M:%S"), "%I%p"),
    hour_display = tolower(sub("^0", "", hour_display))
  ) %>%
  group_by(hour_label, hour_display) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(hour_label)

ggplot(hour_summary, aes(x = hour_label, y = n_incidents)) +
  geom_col(fill = "steelblue") +
  labs(title = "Incidents by Hour of Day",
       x = "Hour", y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Suspension rate by hour of day
ggplot(hour_summary, aes(x = hour_label, y = suspension_rate)) +
  geom_col(fill = "coral") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Hour of Day",
       x = "Hour", y = "Suspension Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r fig.width=11, fig.height=6}
#| warning: false
#| message: false
# Define hours to exclude (midnight through 7am)
# excluded_hours <- sprintf("%02d:00:00", 22:7)

# Heatmap: incidents by day × hour
# dow_hour_table %>%
#   filter(PITS_weekday != "Sunday",
#          !hour_label %in% excluded_hours) %>%
#   mutate(rate_display = ifelse(n_incidents < 5, NA, suspension_rate)) %>%
#   ggplot(aes(x = hour_label, y = PITS_weekday, fill = n_incidents)) +
#   geom_tile(colour = "white") +
#     scale_fill_gradient(low = "white", high = "coral",
#                       name = "Susp. Rate",
#                       labels = scales::percent_format(),
#                       na.value = "grey90") +
#   labs(title = "Incident Count Heatmap: Day of Week × Hour of Day",
#        x = "Hour", y = NULL) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# # Heatmap: suspension rate by day × hour (cells with n < 5 greyed out)
# dow_hour_table %>%
#   filter(PITS_weekday != "Sunday",
#          !hour_label %in% excluded_hours) %>%
#   mutate(rate_display = ifelse(n_incidents < 5, NA, suspension_rate)) %>%
#   ggplot(aes(x = hour_label, y = PITS_weekday, fill = rate_display)) +
#   geom_tile(colour = "white") +
#   scale_fill_gradient(low = "white", high = "coral",
#                       name = "Susp. Rate",
#                       labels = scales::percent_format(),
#                       na.value = "grey90") +
#   labs(title = "Suspension Rate Heatmap: Day of Week × Hour of Day",
#        subtitle = "Grey = fewer than 5 incidents",
#        x = "Hour", y = NULL) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
dow_hour_table <- PITS_deduplicated %>%
  distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(PITS_weekday), !is.na(hour_label)) %>%
  group_by(PITS_weekday, hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

# # Heatmap: incidents by day × hour
ggplot(dow_hour_table,
       aes(x = hour_label, y = PITS_weekday, fill = n_incidents)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = "steelblue",
                      name = "Incidents") +
  labs(title = "Incident Count Heatmap: Day of Week × Hour of Day",
       x = "Hour", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
#
# 
# Heatmap: suspension rate by day × hour (cells with n < 5 greyed out)
# dow_hour_table %>%
#   filter(PITS_weekday != "Sunday") %>%
#   mutate(rate_display = ifelse(n_incidents < 5, NA, suspension_rate)) %>%
#   ggplot(aes(x = hour_label, y = PITS_weekday, fill = rate_display)) +
#   geom_tile(colour = "white") +
#   scale_fill_gradient(low = "white", high = "coral",
#                       name = "Susp. Rate",
#                       labels = scales::percent_format(),
#                       na.value = "grey90") +
#   labs(title = "Suspension Rate Heatmap: Day of Week × Hour of Day (Sunday excluded)",
#        subtitle = "Grey = fewer than 5 incidents",
#        x = "Hour", y = NULL) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

dow_hour_table %>%
  filter(PITS_weekday != "Sunday") %>%
  mutate(
    # Convert PITS_weekday to an ordered factor
    PITS_weekday = factor(PITS_weekday, 
                          levels = c("Saturday", "Friday", "Thursday", "Wednesday",
                                     "Tuesday", "Monday"), 
                          ordered = TRUE),
    rate_display = ifelse(n_incidents < 5, NA, suspension_rate)
  ) %>%
  ggplot(aes(x = hour_label, y = PITS_weekday, fill = rate_display)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = "coral",
                      name = "Susp. Rate",
                      labels = scales::percent_format(),
                      na.value = "grey90") +
  labs(title = "Suspension Rate Heatmap: Day of Week × Hour of Day (Sunday excluded)",
       subtitle = "Grey = fewer than 5 incidents",
       x = "Hour", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# =============================================================================
# SECTION 8: MyPC usage volume by day of week — are busy days incident days?
# =============================================================================
# not statistically significant
```{r}
#| warning: false
#| message: false

# ── 8a. Summarise MyPC usage by day of week ───────────────────────────────────
# MyPC_public <- MyPC_public %>%
#   mutate(MyPC_weekday = factor(MyPC_weekday, levels = weekday_levels))
# 
# mypc_dow <- MyPC_public %>%
#   filter(!is.na(MyPC_weekday)) %>%
#   group_by(MyPC_weekday) %>%
#   summarise(
#     n_sessions    = n(),
#     n_unique_users = n_distinct(new_login_id),
#     .groups = "drop"
#   )

# Generate new_login_id using MD5 hash
# library(digest)
# MyPC_public$new_login_id <- sapply(MyPC_public$`Login ID`, digest, algo="md5")

# # In Section 8a — filter Sunday out of MyPC
# MyPC_dow <- MyPC_public %>%
#   distinct(new_login_id, .keep_all = TRUE) %>%
#   # distinct(`Login ID`, .keep_all = TRUE) %>% 
#   filter(!is.na(MyPC_weekday), MyPC_weekday != "Sunday") %>%
#   group_by(MyPC_weekday) %>%
#   summarise(
#     n_sessions     = n(),
#     n_unique_users = n_distinct(new_login_id),
#     # n_unique_users = n_distinct(`Login ID`),
#     .groups = "drop"
#   )
# 
# MyPC_dow

# ── 8a. Summarise MyPC usage by day of week ───────────────────────────────────
# MyPC_public <- MyPC_public %>%
#   mutate(MyPC_weekday = factor(MyPC_weekday, levels = weekday_levels))
# 
# mypc_dow <- MyPC_public %>%
#   filter(!is.na(MyPC_weekday)) %>%
#   group_by(MyPC_weekday) %>%
#   summarise(
#     n_sessions    = n(),
#     n_unique_users = n_distinct(new_login_id),
#     .groups = "drop"
#   )


```





```{r fig.width = 9, fig.height = 4}
#| warning: false
#| message: false

mypc_dow <- MyPC_public %>%
  distinct(new_login_id, .keep_all = TRUE) %>%
  filter(!is.na(MyPC_weekday), MyPC_weekday != "Sunday") %>%
  group_by(MyPC_weekday) %>%
  summarise(
    n_sessions     = n(),
    n_unique_users = n_distinct(new_login_id),
    .groups = "drop"
  )

# excluded sundays 
dow_combined <- dow_summary_exc %>%
  left_join(mypc_dow, by = c("PITS_weekday" = "MyPC_weekday")) %>%
  mutate(
    incidents_per_100_sessions = (n_incidents / n_sessions) * 100,
    suspensions_per_100_sessions = (n_suspensions / n_sessions) * 100
  )

# ── 8e. Visualise: sessions vs incidents by weekday ───────────────────────────
dow_combined %>%
  dplyr::select(PITS_weekday, n_incidents, n_sessions) %>%
  pivot_longer(cols = c(n_incidents, n_sessions),
               names_to = "metric", values_to = "count") %>%
  dplyr::mutate(metric = dplyr::recode(metric,
                         n_incidents = "Incidents",
                         n_sessions  = "Computer Sessions")) %>%
  ggplot(aes(x = PITS_weekday, y = count, fill = metric)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Computer Sessions" = "steelblue",
                                "Incidents"         = "coral")) +
  labs(title = "Computer Sessions vs. Incidents by Day of Week",
       x = NULL, y = "Count", fill = NULL) +
  theme_minimal()

# ── 8f. Incidents per 100 sessions — normalised incident rate ─────────────────
dow_combined %>% 
  mutate(
    # Convert PITS_weekday to an ordered factor
    PITS_weekday = factor(PITS_weekday, 
                          levels = c("Monday", "Tuesday","Wednesday",
                                     "Thursday", "Friday","Saturday"), 
                          ordered = TRUE)
  ) %>%
ggplot(
       aes(x = PITS_weekday, y = incidents_per_100_sessions)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(incidents_per_100_sessions, 2)),
            vjust = -0.4, size = 3.5) +
  labs(title = "Incidents per 100 Computer Sessions by Day of Week (Sunday excluded)",
       # subtitle = "Normalises for usage volume — higher = more incidents relative to visitors",
       x = NULL, y = "Incidents per 100 Sessions") +
  theme_minimal()

# ── 8g. Suspension rate vs. session volume scatterplot ────────────────────────
# ggplot(dow_combined,
#        aes(x = n_sessions, y = suspension_rate, label = PITS_weekday)) +
#   geom_point(size = 3, colour = "steelblue") +
#   geom_text(vjust = -0.7, size = 3.5) +
#   geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) +
#   scale_y_continuous(labels = scales::percent_format()) +
#   labs(title = "Suspension Rate vs. Computer Session Volume by Weekday (Sunday excluded)",
#        subtitle = "If volume drove suspension rates, points would trend upward",
#        x = "Number of Computer Sessions", y = "Suspension Rate") +
#   theme_minimal()
```



```{r}
#| warning: false
#| message: false

library(ggrepel)
# ── 8g. Suspension rate vs. session volume scatterplot ────────────────────────

ggplot(dow_combined, aes(x = n_sessions, y = suspension_rate, label = PITS_weekday)) + 
  geom_point(size = 3, colour = "steelblue") + 
  geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) + 
  # Added min.segment.length = Inf to hide the dashes
  geom_text_repel(size = 3.5, nudge_y = -0.005, box.padding = 0.4, point.padding = 0.2, min.segment.length = Inf) + 
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(title = "Suspension Rate vs. Computer Session Volume by Weekday (Sunday excluded)", 
       x = "Number of Computer Sessions", 
       y = "Suspension Rate") + 
  theme_minimal()
```


```{r}
# ── 8g. Incidents vs. session volume scatterplot ────────────────────────
# no sundays 
# excluded hours

ggplot(dow_combined, aes(x = n_sessions, y = n_incidents, label = PITS_weekday)) + 
  geom_point(size = 3, colour = "steelblue") + 
  geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) + 
  # Added min.segment.length = Inf to hide the dashes
  geom_text_repel(size = 3.5, nudge_y = -0.005, box.padding = 0.4, point.padding = 0.2, min.segment.length = Inf) + 
  # scale_y_continuous(labels = scales::percent_format()) + 
  labs(title = "Incidents vs. Computer Session Volume by Weekday (Sunday excluded)", 
       x = "Number of Computer Sessions", 
       y = "Incidents") + 
  theme_minimal()
```

produce for hour label 
suspension rate 
```{r}

```


produce for hour label 
incident 
```{r}
# ── Build hourly usage summary from MyPC ──────────────────────────────────────
mypc_hour <- MyPC_public %>%
  distinct(new_login_id, .keep_all = TRUE) %>%
  filter(!is.na(hour_label)) %>%
  group_by(hour_label) %>%
  summarise(
    n_sessions     = n(),
    n_unique_users = n_distinct(new_login_id),
    .groups = "drop"
  )

# ── Join with your existing hour_summary (incidents/suspensions by hour) ──────
hour_combined <- hour_summary %>%
  left_join(mypc_hour, by = "hour_label") %>%
  mutate(
    incidents_per_100_sessions   = (n_incidents / n_sessions) * 100,
    suspensions_per_100_sessions = (n_suspensions / n_sessions) * 100
  )

print(hour_combined)


# ── Filter to top 10 hours by incident count ───────────────────────────────────
library(lubridate)
# hour_top10 <- hour_combined %>%
#   arrange(desc(n_incidents)) %>%
#   slice_head(n = 10) %>%
#   mutate(
#     hour_display = format(strptime(hour_label, format = "%H:%M:%S"), "%-I%p"),
#     hour_display = tolower(hour_display)   # "4pm" instead of "4PM"
#   )
# 
# # ── Scatterplot: sessions vs incidents, labeled by hour ────────────────────────
# ggplot(hour_top10, aes(x = n_sessions, y = n_incidents, label = hour_display)) +
#   geom_point(size = 3, colour = "steelblue") +
#   geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) +
#   geom_text_repel(size = 3.5, nudge_y = 1, box.padding = 0.4,
#                   point.padding = 0.2, min.segment.length = Inf) +
#   labs(title = "Incidents vs. Computer Session Volume — Top 10 Hours by Incident Count",
#        x = "Number of Computer Sessions", y = "Incidents") +
#   theme_minimal()

# ── Filter to top 10 hours by incident count ───────────────────────────────────
hour_top10 <- hour_combined %>%
  arrange(desc(n_incidents)) %>%
  slice_head(n = 10)

# ── Scatterplot: sessions vs incidents, labeled by hour ────────────────────────
ggplot(hour_top10, aes(x = n_sessions, y = n_incidents, label = hour_label)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, colour = "coral", linewidth = 0.8) +
  geom_text_repel(size = 3.5, nudge_y = 1, box.padding = 0.4,
                  point.padding = 0.2, min.segment.length = Inf) +
  labs(title = "Incidents vs. Computer Session Volume — Top 10 Hours by Incident Count",
       x = "Number of Computer Sessions", y = "Incidents") +
  theme_minimal()
```


4.  Mixed-effects model — treat location as a random effect
a mixed model quantifies how much of the unexplained variance in suspension is attributable to branch-level clustering, independent of time effects.
```{r}
# Hours to exclude: 21:00–23:00 and 00:00–09:00 (keep only 9:00–20:00)
excluded_hours <- sprintf("%02d:00:00", c(21:23, 0:9))

# levels(PITS_deduplicated$hour_label)
# excluded_hours %in% levels(PITS_deduplicated$hour_label)

combined_df_exc <- PITS_deduplicated %>%
  filter(PITS_weekday != "Sunday") %>% 
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label),
         !hour_label %in% excluded_hours) %>%
  mutate(hour_label = droplevels(hour_label))  # drop the now-unused factor levels

library(lme4)

mixed_model <- glmer(
  susp_binary ~ 
    PITS_weekday +
    hour_label + 
    (1 | Location_masked),
  data   = combined_df_exc,
  family = binomial
)
summary(mixed_model)

# Intraclass correlation — how much variance is due to branch?
library(performance)
icc(mixed_model)
```


deviations from the average log-odds of suspension
```{r}
# library(lme4)

ranef_branch <- ranef(mixed_model)$Location_masked
ranef_branch$Location_masked <- rownames(ranef_branch)
names(ranef_branch)[1] <- "random_effect"

ranef_branch <- ranef_branch %>%
  arrange(desc(random_effect))

print(ranef_branch)

# Visualize
ggplot(ranef_branch, aes(x = reorder(Location_masked, random_effect),
                         y = random_effect)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Branch-Level Random Effects (Suspension Likelihood)",
       subtitle = "excluding Sundays and hours 8pm to 9am",
       x = NULL, y = "Random Effect (log-odds deviation)") +
  theme_minimal()
```
```{r}

# 1. Fit your mixed model
model <- lmer(Suspension_length ~ (1 | Location_masked), data = PITS_deduplicated)

# 2. Extract variance components
variance_components <- as.data.frame(VarCorr(model))

# 3. Pull out specific variances
var_between  <- variance_components$vcov[variance_components$grp == "Location_masked"]
var_residual <- variance_components$vcov[variance_components$grp == "Residual"]

# 4. Calculate the equation
manual_icc <- var_between / (var_between + var_residual)

cat("Manual Intercept ICC:", manual_icc, "\n")
```



```{r fig.width=10, fig.height=4}
# ── 9h. Visualise suspension rate by month ────────────────────────────────────
ggplot(month_summary, aes(x = month_name, y = suspension_rate, group = 1)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = scales::percent(suspension_rate, accuracy = 1)),
            vjust = -0.4, size = 3.2) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Month",
       x = NULL, y = "Suspension Rate") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Season comparison
season_summary <- PITS_deduplicated %>%
  filter(!is.na(season)) %>%
  group_by(season) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(season_summary, aes(x = season, y = suspension_rate)) +
  geom_col(fill = "coral") +
  geom_text(aes(label = scales::percent(suspension_rate, accuracy = 1)),
            vjust = -0.4, size = 3.5) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Suspension Rate by Season",
       x = NULL, y = "Suspension Rate") +
  theme_minimal()
```




```{r fig.width=8, fig.height=6}
# Prepare data — ensure season, weekday, hour, and location are all present
mixed_df <- PITS_deduplicated %>%
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label), !is.na(season)) %>% 
  filter(PITS_weekday != "Sunday",
         !hour_label %in% excluded_hours)

# ── Mixed model: weekday + hour + season as fixed effects, branch as random ──
mixed_model_season <- glmer(
  susp_binary ~ PITS_weekday + hour_label + season + (1 | Location_masked),
  data    = mixed_df,
  family  = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(mixed_model_season)

# ── ICC — how much variance is due to branch, controlling for weekday/hour/season ──
icc_season <- icc(mixed_model_season)
print(icc_season)

# ── Extract and rank branch-level random effects (with season controlled) ────
ranef_branch_season <- ranef(mixed_model_season)$Location_masked
ranef_branch_season$Location_masked <- rownames(ranef_branch_season)
names(ranef_branch_season)[1] <- "random_effect"

ranef_branch_season <- ranef_branch_season %>%
  arrange(desc(random_effect))


ggplot(ranef_branch_season, aes(x = reorder(Location_masked, random_effect),
                                y = random_effect)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(title = "Branch-Level Random Effects (Suspension Likelihood)",
       subtitle = "Controlling for weekday, hour, and season",
       x = NULL, y = "Random Effect (log-odds deviation)") +
  theme_minimal()
```
