------------------------------------------------------------------------
title: "PITS_statistical_tests" 
format: html 
editor: visual 
---

### read in libraries

```{r}
library(readr)
library(tidyverse)
library(broom)       # tidy model output
library(MASS)        # glm.nb for negative binomial
library(dunn.test)   # Dunn post-hoc after Kruskal-Wallis
```

### read in data
```{r}
PITS_public <- read_csv("processed_data/PITS_public_share.csv") 
# %>% select(-1)
MyPC_public  <- read_csv("processed_data/MyPC_public_share.csv") 
# %>% select(-1)

```

# =============================================================================
# Clean PITS
# =============================================================================
```{r}
PITS_public <- PITS_public %>%
  mutate(
    # Convert string "NA" to real NA
    `Suspension Status` = na_if(`Suspension Status`, "NA"),
    `Suspension Type`   = na_if(`Suspension Type`,   "NA"),
    Susp_Start_date     = na_if(as.character(Susp_Start_date), "NA"),
    Susp_End_date       = na_if(as.character(Susp_End_date),   "NA"),
    Susp_Start_date     = as.Date(Susp_Start_date),
    Susp_End_date       = as.Date(Susp_End_date),
    Date2               = as.Date(Date2),

    Suspension_length = case_when(
      is.na(Susp_End_date) | is.na(Susp_Start_date) ~ NA_real_,
      Susp_End_date < Susp_Start_date               ~ NA_real_,
      TRUE ~ as.numeric(Susp_End_date - Susp_Start_date)
    ),

    # Drive susp_binary off Suspension Type — avoids "NONE" miscount
    susp_binary = ifelse(`Suspension Type` == "Suspended customer", 1L, 0L),
    susp_binary = replace_na(susp_binary, 0L),

    hour_label = fct_reorder(as.factor(hour_label), hour_label)
  )
PITS_public$PITS_weekday <- weekdays(PITS_public$Date2)

```

MyPC data
```{r}
MyPC_public <- MyPC_public %>%
  mutate(
    booking_date = as.Date(booking_date),
    booking_from = as.Date(booking_from),
    hour_label   = fct_reorder(as.factor(hour_label), hour_label)
  )
MyPC_public$MyPC_weekday <- weekdays(MyPC_public$booking_date)

```


# =============================================================================
# Pre-compute ratio data (computer users vs incidents)
# =============================================================================
```{r}
PITS_incident_counts <- PITS_public %>%
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(incident_count = n_distinct(`Incident ID`), .groups = "drop")

MyPC_counts <- MyPC_public %>%
  group_by(Location_masked, date = booking_date) %>%
  summarise(user_count = n(), .groups = "drop")

ratio_data <- MyPC_counts %>%
  full_join(PITS_incident_counts, by = c("Location_masked", "date")) %>%
  mutate(
    user_count     = replace_na(user_count,     0),
    incident_count = replace_na(incident_count, 0)
  )

```


# =============================================================================
# Pre-compute suspension ratio data (suspensions per incident + avg length)
# =============================================================================
```{r}

# One row per unique suspended incident — prevents duration double-counting
PITS_susp_deduped <- PITS_public %>%
  filter(susp_binary == 1,
         !is.na(`Incident ID`),
         !is.na(Suspension_length)) %>%
  distinct(`Incident ID`, Location_masked, Date2, Suspension_length)

# Suspensions per location per day
PITS_susp_counts <- PITS_susp_deduped %>%
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(
    suspension_count   = n_distinct(`Incident ID`),
    avg_susp_length    = mean(Suspension_length),
    median_susp_length = median(Suspension_length),
    .groups = "drop"
  )

susp_ratio_data <- PITS_incident_counts %>%
  left_join(PITS_susp_counts, by = c("Location_masked", "date")) %>%
  mutate(
    suspension_count   = replace_na(suspension_count,   0),
    avg_susp_length    = replace_na(avg_susp_length,    0),
    median_susp_length = replace_na(median_susp_length, 0),
    susp_ratio         = ifelse(incident_count > 0,
                                suspension_count / incident_count, 0)
  )

```



# =============================================================================
# SECTION 1: Baseline — Suspensions as the dominant outcome
# =============================================================================

## Descriptive: suspension rate overall and by location
```{r}
suspension_summary <- PITS_public %>%
  summarise(
    total_incidents    = n_distinct(`Incident ID`),
    total_suspensions  = sum(susp_binary, na.rm = TRUE),
    suspension_rate    = mean(susp_binary, na.rm = TRUE)
  )

suspension_summary_by_loc <- PITS_public %>%
  group_by(Location_masked) %>%
  summarise(
    total_incidents   = n_distinct(`Incident ID`),
    total_suspensions = sum(susp_binary, na.rm = TRUE),
    suspension_rate   = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(suspension_rate))

print(suspension_summary)
print(suspension_summary_by_loc)
```

## One-sample proportion test: Is suspension rate > 50%?
```{r}
# H0: true suspension rate = 0.50
# H1: true suspension rate > 0.50
binom_test <- binom.test(
  x = suspension_summary$total_suspensions,
  n = suspension_summary$total_incidents,
  p = 0.50,
  alternative = "greater"
)
print(binom_test)
```

## Support service use vs. suspension rate
```{r}
support_summary <- PITS_public %>%
  summarise(
    n                  = n(),
    pct_suspended      = mean(susp_binary, na.rm = TRUE),
    pct_guard_called   = mean(`Support:Called Guard?`         == "Yes", na.rm = TRUE),
    pct_police_called  = mean(`Support:Called Police (911)?`  == "Yes", na.rm = TRUE),
    pct_emt_called     = mean(`Support:Called EMT?`           == "Yes", na.rm = TRUE),
    pct_crisis_called  = mean(`Support:Called crisis line?`   == "Yes", na.rm = TRUE)
  ) %>%
  pivot_longer(-n, names_to = "response_type", values_to = "rate") %>%
  mutate(response_type = gsub("pct_", "", response_type))

print(support_summary)

# McNemar test: suspension vs. guard called (paired binary comparison)
# Requires a contingency table of the two binary variables
mcnemar_df <- PITS_public %>%
  mutate(
    guard_called = ifelse(`Support:Called Guard?` == "Yes", 1L, 0L),
    guard_called = replace_na(guard_called, 0L)
  )

mcnemar_table <- table(
  suspended    = mcnemar_df$susp_binary,
  guard_called = mcnemar_df$guard_called
)

print(mcnemar_table)
mcnemar_result <- mcnemar.test(mcnemar_table)
print(mcnemar_result)
```


# =============================================================================
# SECTION 2 (Main): Logistic regression — does reporting predict suspension
#                   better than incident characteristics do?
# =============================================================================

```{r, echo=FALSE}
# Prepare: drop rows missing key predictors
pits_model_df <- PITS_public %>%
  filter(!is.na(susp_binary),
         !is.na(parsed_categories),
         !is.na(hour_label),
         !is.na(Location_masked),
         !is.na(PITS_weekday)) %>%
  mutate(
    infraction_cat = as.factor(parsed_categories),
    location       = as.factor(Location_masked),
    weekday        = as.factor(PITS_weekday),
    hour           = as.factor(hour_label)
  )

# Null model (intercept only — "reporting alone")
logit_null <- glm(susp_binary ~ 1,
                  data   = pits_model_df,
                  family = binomial)

# Full model with incident characteristics
logit_full <- glm(susp_binary ~ infraction_cat + hour + weekday + location,
                  data   = pits_model_df,
                  family = binomial)

summary(logit_full)

# McFadden pseudo-R² — low value supports argument that
# observable characteristics don't explain suspension well
mcfadden_r2 <- 1 - (as.numeric(logLik(logit_full)) /
                      as.numeric(logLik(logit_null)))
cat("McFadden pseudo-R²:", round(mcfadden_r2, 4), "\n")

# Tidy coefficient table
logit_tidy <- tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
  arrange(p.value)
print(logit_tidy)

# Likelihood ratio test: does the full model fit better than null?
lrt_result <- anova(logit_null, logit_full, test = "LRT")
print(lrt_result)
```


# =============================================================================
# SECTION 3: Chi-square — suspensions not proportional to infraction type
# =============================================================================
A non-significant result here supports you: suspensions are equally likely regardless of what the infraction was
```{r}
# Contingency table: infraction category × suspension binary
infraction_susp_table <- table(
  infraction = PITS_public$parsed_categories,
  suspended  = PITS_public$susp_binary
)

print(infraction_susp_table)

chi_infraction <- chisq.test(infraction_susp_table)
print(chi_infraction)

# Expected vs observed suspension rate per infraction category
infraction_rates <- PITS_public %>%
  filter(!is.na(parsed_categories)) %>%
  group_by(parsed_categories) %>%
  summarise(
    n               = n(),
    n_suspended     = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(suspension_rate))

print(infraction_rates)
```


# =============================================================================
# SECTION 4: Kruskal-Wallis — suspension length not calibrated to infraction
# =============================================================================
Tests whether suspension length varies by infraction category — if it doesn't, lengths aren't being calibrated to severity either
```{r}
# Filter to suspended incidents only, with valid length
susp_length_df <- PITS_public %>%
  filter(susp_binary == 1,
         !is.na(Suspension_length),
         !is.na(parsed_categories))

# Descriptive: median length by infraction category
length_by_infraction <- susp_length_df %>%
  group_by(parsed_categories) %>%
  summarise(
    n              = n(),
    median_days    = median(Suspension_length),
    mean_days      = mean(Suspension_length),
    sd_days        = sd(Suspension_length),
    .groups = "drop"
  ) %>%
  arrange(desc(median_days))

print(length_by_infraction)

# Kruskal-Wallis test
kw_length <- kruskal.test(
  Suspension_length ~ as.factor(parsed_categories),
  data = susp_length_df
)
print(kw_length)

# Dunn post-hoc test (only meaningful if KW is significant,
# but run regardless — non-significant pairs support your argument)
dunn_length <- dunn.test(
  x     = susp_length_df$Suspension_length,
  g     = susp_length_df$parsed_categories,
  method = "bonferroni",
  kw    = TRUE
)

# Kruskal-Wallis by location
kw_location <- kruskal.test(
  Suspension_length ~ as.factor(Location_masked),
  data = susp_length_df
)
print(kw_location)
```


# =============================================================================
# SECTION 5 (Supporting): Poisson/NB regression — usage volume does not
#                          predict incident counts
# =============================================================================

```{r}
# Poisson model: do more computer users → more incidents?
poisson_model <- glm(
  incident_count ~ user_count + Location_masked,
  data   = ratio_data,
  family = poisson
)
summary(poisson_model)

# Check for overdispersion (variance >> mean → use negative binomial)
dispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) /
                    poisson_model$df.residual
cat("Dispersion ratio:", round(dispersion_ratio, 3),
    " (>1.5 suggests overdispersion → prefer NB)\n")

# Negative binomial model (more robust if overdispersed)
nb_model <- glm.nb(
  incident_count ~ user_count + Location_masked,
  data = ratio_data
)
summary(nb_model)

# Tidy both for comparison
tidy_poisson <- tidy(poisson_model, exponentiate = TRUE, conf.int = TRUE)
tidy_nb      <- tidy(nb_model,      exponentiate = TRUE, conf.int = TRUE)

cat("\n--- Poisson IRRs ---\n"); print(tidy_poisson)
cat("\n--- Negative Binomial IRRs ---\n"); print(tidy_nb)
```


# =============================================================================
# SECTION 6: Chi-square — suspension rates vary by location (inconsistent
#            application)
# =============================================================================

```{r}
# Contingency table: location × suspension binary
location_susp_table <- table(
  location  = PITS_public$Location_masked,
  suspended = PITS_public$susp_binary
)

print(location_susp_table)

chi_location <- chisq.test(location_susp_table)
print(chi_location)

# Standardised residuals — which locations suspend more/less than expected?
std_residuals <- as.data.frame(chi_location$stdres) %>%
  setNames(c("location", "suspended", "std_resid")) %>%
  filter(suspended == "1") %>%
  arrange(desc(std_resid))

cat("\nStandardised residuals (suspended == 1):\n")
print(std_residuals)

# Logistic regression: location as predictor of suspension
logit_location <- glm(
  susp_binary ~ Location_masked,
  data   = PITS_public,
  family = binomial
)
summary(logit_location)
tidy(logit_location, exponentiate = TRUE, conf.int = TRUE)
```

