------------------------------------------------------------------------
title: "PITS_statistical_tests" 
format: html 
editor: visual 
---

### read in libraries
last run 2026-07-31

```{r}
library(readr)
library(lubridate)
library(rstatix)
library(tidyverse)
library(broom)       # tidy model output
library(MASS)        # glm.nb for negative binomial
library(dunn.test)   # Dunn post-hoc after Kruskal-Wallis
library(dplyr)
library(purrr)
library(performance)
```

### read in data
```{r}
PITS_public <- read_csv("processed_data/PITS_public_share.csv") 
# %>% select(-1)
MyPC_public  <- read_csv("processed_data/MyPC_public_share.csv") 
# %>% select(-1)

```
Share data for pits public is up to 5034 rows with duplicated columns for infractions

### Drop any extraneous columns
```{r}
# MyPC_public <- MyPC_public %>% 
#   dplyr::select(-Site, -Resource)

```


# ========================================================================
# Clean PITS
# ========================================================================
```{r}
PITS_public <- PITS_public %>%
  mutate(
    # Convert string "NA" to real NA
    `Suspension Status` = na_if(`Suspension Status`, "NA"),
    `Suspension Type`   = na_if(`Suspension Type`,   "NA"),
    Susp_Start_date     = na_if(as.character(Susp_Start_date), "NA"),
    Susp_End_date       = na_if(as.character(Susp_End_date),   "NA"),
    Susp_Start_date     = as.Date(Susp_Start_date),
    Susp_End_date       = as.Date(Susp_End_date),
    Date2               = as.Date(Date2),

    Suspension_length = case_when(
      is.na(Susp_End_date) | is.na(Susp_Start_date) ~ NA_real_,
      Susp_End_date < Susp_Start_date               ~ NA_real_,
      TRUE ~ as.numeric(Susp_End_date - Susp_Start_date)
    ),

    # Drive susp_binary off Suspension Type — avoids "NONE" miscount
    susp_binary = ifelse(`Suspension Type` == "Suspended customer", 1L, 0L),
    susp_binary = replace_na(susp_binary, 0L),

    hour_label = fct_reorder(as.factor(hour_label), hour_label)
  )
PITS_public$PITS_weekday <- weekdays(PITS_public$Date2)

```

### Season 
```{r}
PITS_public <- PITS_public %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  mutate(
    month_num   = as.integer(format(Date2, "%m")),
    month_name  = factor(format(Date2, "%B"),
                         levels = month.name),  # orders Jan–Dec correctly
    season      = case_when(
      month_num %in% c(6, 7, 8)  ~ "Summer",
      month_num %in% c(9, 10, 11) ~ "Fall",
      month_num %in% c(12, 1, 2)  ~ "Winter",
      month_num %in% c(3, 4, 5)   ~ "Spring",
      TRUE ~ NA_character_
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )

```


MyPC data
```{r}
MyPC_public <- MyPC_public %>%
  mutate(
    booking_date = as.Date(booking_date),
    booking_from = as.Date(booking_from),
    hour_label   = fct_reorder(as.factor(hour_label), hour_label)
  )
MyPC_public$MyPC_weekday <- weekdays(MyPC_public$booking_date)

```


# =============================================================================
# Pre-compute ratio data (computer users vs incidents)
# =============================================================================
```{r}
# this is just to get a ratio and does not need the hierarchical categories
PITS_incident_counts <- PITS_public %>%
  distinct(`Incident ID`, .keep_all = TRUE) %>% 
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(incident_count = n_distinct(`Incident ID`), .groups = "drop")

MyPC_counts <- MyPC_public %>%
  distinct(new_login_id, .keep_all = TRUE) %>% 
  group_by(Location_masked, date = booking_date) %>%
  summarise(user_count = n(), .groups = "drop")

ratio_data <- MyPC_counts %>%
  full_join(PITS_incident_counts, by = c("Location_masked", "date")) %>%
  mutate(
    user_count     = replace_na(user_count,     0),
    incident_count = replace_na(incident_count, 0)
  )

```


# =============================================================================
# Pre-compute suspension ratio data (suspensions per incident + avg length)
# =============================================================================
```{r}

# One row per unique suspended incident — prevents duration double-counting
PITS_susp_deduped <- PITS_public %>%
  distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(susp_binary == 1,
         !is.na(`Incident ID`),
         !is.na(Suspension_length)) %>%
  distinct(`Incident ID`, Location_masked, Date2, Suspension_length)

# Suspensions per location per day
PITS_susp_counts <- PITS_susp_deduped %>%
  group_by(Location_masked, date = as.Date(Date2)) %>%
  summarise(
    suspension_count   = n_distinct(`Incident ID`),
    avg_susp_length    = mean(Suspension_length),
    median_susp_length = median(Suspension_length),
    .groups = "drop"
  )
# again severity will come into play later
susp_ratio_data <- PITS_incident_counts %>%
  left_join(PITS_susp_counts, by = c("Location_masked", "date")) %>%
  mutate(
    suspension_count   = replace_na(suspension_count,   0),
    avg_susp_length    = replace_na(avg_susp_length,    0),
    median_susp_length = replace_na(median_susp_length, 0),
    susp_ratio         = ifelse(incident_count > 0,
                                suspension_count / incident_count, 0)
  )

```



# =============================================================================
# SECTION 1: Baseline — Suspensions as the dominant outcome
# =============================================================================

```{r}
# unique_count <- n_distinct(PITS_public$`Incident ID`) #2995
```

## Descriptive: suspension rate overall and by location
```{r}
suspension_summary <- PITS_public %>%
  distinct(`Incident ID`, susp_binary) %>%
  summarise(
    total_incidents    = n_distinct(`Incident ID`),
    total_suspensions  = sum(susp_binary, na.rm = TRUE),
    suspension_rate    = mean(susp_binary, na.rm = TRUE)
  )


suspension_summary_by_loc <- PITS_public %>%
  group_by(Location_masked) %>%
  distinct(`Incident ID`, susp_binary) %>%
  summarise(
    total_incidents   = n_distinct(`Incident ID`),
    total_suspensions = sum(susp_binary, na.rm = TRUE),
    suspension_rate   = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(suspension_rate))

print(suspension_summary)
print(suspension_summary_by_loc)
```

## One-sample proportion test: Is suspension rate > 50%?
```{r}
# H0: true suspension rate = 0.50
# H1: true suspension rate > 0.50
binom_test <- binom.test(
  x = suspension_summary$total_suspensions,
  n = suspension_summary$total_incidents,
  p = 0.50,
  alternative = "greater"
)
binom_test
```

# =============================================================================
# SECTION 2: predictor cramers v, pseudo r^2, K-W, dunn
# =============================================================================


# establish deduplicated dataset
```{r}
library(dplyr)
library(purrr)
library(rstatix)
library(performance)

# 1. Establish your clean, deduplicated baseline data (One row per Incident ID)
severity_hierarchy <- c(
  "Dangerous, Threatening, or Illegal Conduct", 
  "Disruptive Personal Behavior", 
  "Use and preservation of library materials and property", 
  "Incident Impact Level - Staff Traumatic Stress", 
  "Other", 
  "Children in the Library", 
  "No Infractions Set", 
  "Medical 911"
)



# Create an independent, deduplicated dataset using your severity logic
PITS_deduplicated <- PITS_public %>% 
  filter(
    !is.na(parsed_categories),
    parsed_categories %in% severity_hierarchy
  ) %>% 
  mutate(parsed_categories = factor(parsed_categories, 
                                    levels = severity_hierarchy, 
                                    ordered = TRUE)) %>% 
  arrange(`Incident ID`, parsed_categories) %>% 
  distinct(`Incident ID`, .keep_all = TRUE) # Now each incident counts exactly once

# PITS_deduplicated <- PITS_deduplicated %>%
#   mutate(
#     # Automatically lumps categories with < 30 total rows into the custom name
#     parsed_categories = fct_lump_min(
#       as.factor(parsed_categories), 
#       min = 30, 
#       other_level = "Minor Infractions"
#     ),
#     parsed_categories = as.character(parsed_categories)
#   )
```

### create minor infractions category
```{r}
library(dplyr)
library(forcats)


PITS_deduplicated <- PITS_deduplicated %>%
  mutate(
    parsed_categories1 = as.character(parsed_categories),
    parsed_categories1 = case_when(
      parsed_categories1 %in% c(
        "Incident Impact Level - Staff Traumatic Stress", 
        "No Infractions Set", 
        "Medical", 
        "Children in the Library"
      ) ~ "Minor Infractions",
      TRUE ~ parsed_categories
    )
  )

```


```{r}
# # 1. Isolate the filtered data for this specific predictor
check_df <- PITS_deduplicated %>%
  filter(!is.na(susp_binary), !is.na(parsed_categories1))

# 2. Check the number of unique categories remaining
print(paste("Unique outcome levels:", length(unique(check_df$susp_binary))))
print(paste("Unique predictor levels:", length(unique(check_df$parsed_categories1))))

# 3. View the raw contingency table
raw_table <- table(as.factor(check_df$susp_binary), as.factor(check_df$parsed_categories1))
print(raw_table)
```

# Predictor Cramers_V Pseudo_R2_Nagelkerke
```{r}
# use shortened parsed_categories1 which has the minor infractions category
# 2. Define your modeling elements
outcome_var <- "susp_binary"
predictors  <- c("parsed_categories1", "Location_masked", "hour_label", 
                 "PITS_weekday", "season", "month_name")

# 3. Map through predictors using the deduplicated data
metrics_table <- map_df(predictors, function(pred) {
  
  # Clean missing values using the deduplicated baseline
  clean_df <- PITS_deduplicated %>% 
    filter(!is.na(.data[[outcome_var]]), !is.na(.data[[pred]]))
  
  # Create the raw contingency table
  tab <- table(as.factor(clean_df[[outcome_var]]), as.factor(clean_df[[pred]]))

  # # Calculate Cramér's V safely
  v_val <- tryCatch({
  if (ncol(tab) < 2 || nrow(tab) < 2) {
    0
  } else {
    # Run a standard base R chi-square test
    chi_test <- chisq.test(tab, correct = TRUE)
    chi_sq <- chi_test$statistic
    n_total <- sum(tab)
    
    # Mathematical definition of Cramer's V
    min_dim <- min(nrow(tab) - 1, ncol(tab) - 1)
    
    res <- sqrt(chi_sq / (n_total * min_dim))
    
    # Strip names and handle any unexpected NaN/NA values safely
    res <- as.numeric(res)
    if (is.na(res) || is.nan(res)) 0 else res
  }
}, error = function(e) {
  return(0) # Returns 0 instead of killing the column with NA
})
#   # Tschuprow's T 
#   v_val <- tryCatch({
#   if (ncol(tab) < 2 || nrow(tab) < 2) {
#     0
#   } else {
#     # Tschuprow's T handles your 2xN sparse table structure cleanly
#     res <- vcd::assocstats(tab)$tschuprow
#     if (is.nan(res) || is.na(res)) 0 else res
#   }
# }, error = function(e) NA)
  # Calculate Nagelkerke Pseudo R-squared
  model_formula <- as.formula(paste(outcome_var, "~", pred))
  fit <- glm(model_formula, data = clean_df, family = binomial(link = "logit"))
  
  pseudo_r2 <- tryCatch({
    as.numeric(performance::r2_nagelkerke(fit))
  }, error = function(e) NA)
  
  # Return data frame row
  data.frame(
    Predictor = pred,
    Cramers_V = round(v_val, 4),
    Pseudo_R2_Nagelkerke = round(pseudo_r2, 4)
  )
})

# 4. results table
# View your accurate metrics table
metrics_table <- metrics_table %>% arrange(desc(Pseudo_R2_Nagelkerke))
metrics_table
```

# descriptive stats on whether other interventions affect suspension
```{r}
#| warning: false
#| message: false

## Support service use vs. suspension rate

library(dplyr)
library(tidyr)

support_summary <- PITS_deduplicated %>% 
  distinct(`Incident ID`, .keep_all = TRUE) %>% 
  summarise(
    # 1. Calculate raw counts (n)
    n_suspended = sum(susp_binary == 1, na.rm = TRUE),
    n_guard_called = sum(`Support:Called Guard?` == "Yes", na.rm = TRUE),
    n_police_called = sum(`Support:Called Police (911)?` == "Yes", na.rm = TRUE),
    n_emt_called = sum(`Support:Called EMT?` == "Yes", na.rm = TRUE),
    n_crisis_called = sum(`Support:Called crisis line?` == "Yes", na.rm = TRUE),
    
    # 2. Calculate rates
    pct_suspended = mean(susp_binary, na.rm = TRUE),
    pct_guard_called = mean(`Support:Called Guard?` == "Yes", na.rm = TRUE),
    pct_police_called = mean(`Support:Called Police (911)?` == "Yes", na.rm = TRUE),
    pct_emt_called = mean(`Support:Called EMT?` == "Yes", na.rm = TRUE),
    pct_crisis_called = mean(`Support:Called crisis line?` == "Yes", na.rm = TRUE)
  ) %>% 
  # 3. Pivot both sets of columns at the same time
  pivot_longer(
    cols = everything(),
    names_to = c(".value", "response_type"),
    names_pattern = "(n|pct)_(.*)"
  )

support_summary

# McNemar test: suspension vs. guard called (paired binary comparison)
# Requires a contingency table of the two binary variables
mcnemar_df <- PITS_public %>%
   distinct(`Incident ID`, .keep_all = TRUE) %>% 
  mutate(
    guard_called = ifelse(`Support:Called Guard?` == "Yes", 1L, 0L),
    guard_called = replace_na(guard_called, 0L)
  )

mcnemar_table <- table(
  suspended    = mcnemar_df$susp_binary,
  guard_called = mcnemar_df$guard_called
)

print(mcnemar_table)
mcnemar_result <- mcnemar.test(mcnemar_table)
print(mcnemar_result)
```




## Checking model fit and multi-collinearity


 =============================================================================
 SECTION 2 (Main): Logistic regression — does reporting predict suspension
                   better than incident characteristics do?
 =============================================================================

```{r, echo=FALSE}
#| warning: false
#| message: false

# weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday",
#                     "Friday", "Saturday", "Sunday")
# 
# PITS_public <- PITS_public %>%
#   mutate(PITS_weekday = factor(PITS_weekday, levels = weekday_levels))
# 
# # Define hours to exclude (midnight through 7am)
# excluded_hours <- sprintf("%02d:00:00", 00:7)

pits_model_df <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>%  
  # filter(PITS_weekday != "Sunday",
  #       !hour_label %in% excluded_hours) %>%
  filter(!is.na(susp_binary),
         !is.na(parsed_categories1),
         !is.na(hour_label),
         !is.na(Location_masked),
         # !is.na(month_name),
         !is.na(PITS_weekday)) %>%
  mutate(
    infraction_cat = as.factor(parsed_categories1),
    location       = as.factor(Location_masked),
    weekday        = as.factor(PITS_weekday),
    hour           = as.factor(hour_label)
  )

# Null model (intercept only — "reporting alone")
logit_null <- glm(susp_binary ~ 1,
                  data   = pits_model_df,
                  family = binomial)

# Full model with incident characteristics
logit_full <- glm(susp_binary ~ infraction_cat + hour + weekday + location,
                  data   = pits_model_df,
                  family = binomial)
# 
summary(logit_full)

# McFadden pseudo-R² — low value supports argument that
# observable characteristics don't explain suspension well
mcfadden_r2 <- 1 - (as.numeric(logLik(logit_full)) /
                      as.numeric(logLik(logit_null)))
cat("McFadden pseudo-R²:", round(mcfadden_r2, 4), "\n")
# 
# # Tidy coefficient table
logit_tidy <- tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
   arrange(p.value)
 print(logit_tidy)

#
# Likelihood ratio test: does the full model fit better than null?

lrt_result <- anova(logit_null, logit_full, test = "LRT")

print(lrt_result)

```

Different Logit model with just location
```{r}
# Full model with incident characteristics
logit_loc_hour <- glm(susp_binary ~  location + hour,
                  data   = pits_model_df,
                  family = binomial)

summary(logit_loc_hour)

# McFadden pseudo-R² — low value supports argument that
# observable characteristics don't explain suspension well
mcfadden_r2 <- 1 - (as.numeric(logLik(logit_loc_hour)) /
                      as.numeric(logLik(logit_null)))
cat("McFadden pseudo-R²:", round(mcfadden_r2, 4), "\n")

# # Tidy coefficient table
# logit_tidy <- tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
#   arrange(p.value)
# print(logit_tidy)

# Likelihood ratio test: does the full model fit better than null?
lrt_result <- anova(logit_null, logit_loc_hour, test = "LRT")
print(lrt_result)
```


# =============================================================================
# SECTION 3: Chi-square — suspensions not proportional to infraction type
# =============================================================================
changed to deduplicated data
```{r}
#| warning: false
#| message: false

# Contingency table: infraction category × suspension binary
infraction_susp_table <- table(
  infraction = PITS_deduplicated$parsed_categories1,
  suspended  = PITS_deduplicated$susp_binary
)

print(infraction_susp_table)

chi_infraction <- chisq.test(infraction_susp_table)
print(chi_infraction)

# # Expected vs observed suspension rate per infraction category
# infraction_rates <- PITS_deduplicated %>%
#   distinct(`Incident ID`, .keep_all = TRUE) %>%
#   filter(!is.na(parsed_categories1)) %>%
#   group_by(parsed_categories) %>%
#   summarise(
#     n               = n(),
#     n_suspended     = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   arrange(desc(suspension_rate))
# 
# print(infraction_rates)
```


```{r}
# library(dplyr)
# library(gt)
# # 
# # 1. Run your core analysis with string formatting additions
# infraction_table_data <- PITS_deduplicated %>% 
#   filter(!is.na(parsed_categories1)) %>%
#   group_by(parsed_categories1) %>%
#   summarise(
#     n = n(),
#     n_suspended = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   arrange(desc(suspension_rate)) %>%
#   # Apply sprintf to create a clean "n (% sample)" style or isolated percentage
#   mutate(
#     pct_formatted = sprintf("%.1f%%", suspension_rate * 100)
#   )
# 
# # 2. Convert data frame into a publication-ready gt object
# publishable_table <- infraction_table_data %>%
#   select(parsed_categories, n, n_suspended, pct_formatted) %>%
#   gt() %>%
# 
#   # Add clear, academic headers
#   tab_header(
#     title = md("**Suspension Rates Across Infraction Categories**"),
#     subtitle = "Observed incident counts and suspension frequencies"
#   ) %>%
# 
#   # Clean up column labels (replaces ugly variable names)
#   cols_label(
#     parsed_categories = "Infraction Category",
#     n = "Total Incidents (N)",
#     n_suspended = "Suspensions (n)",
#     pct_formatted = "Suspension Rate (%)"
#   ) %>%
# 
#   # Apply professional academic styling rules (Minimalist/APA style)
#   tab_options(
#     table.border.top.color = "black",
#     table.border.top.width = px(2),
#     table.border.bottom.color = "black",
#     table.border.bottom.width = px(2),
#     heading.border.bottom.color = "black",
#     heading.border.bottom.width = px(1),
#     column_labels.border.bottom.color = "black",
#     column_labels.border.bottom.width = px(1.5),
#     table_body.border.bottom.color = "black",
#     table_body.border.bottom.width = px(1),
#     row.striping.background_color = "#f9f9f9",
#     table.background.color = "white"
#   ) %>%
# 
#   # Opt-in to subtle zebra striping for easier scanning
#   opt_row_striping() %>%
# 
#   # Center the numeric data, left-align the categories
#   cols_align(align = "left", columns = parsed_categories) %>%
#   cols_align(align = "center", columns = c(n, n_suspended, pct_formatted))
# infraction_table_data
```


# =============================================================================
# SECTION 4: Kruskal-Wallis — suspension length not calibrated to infraction
# =============================================================================
```{r}
#| warning: false
#| message: false

# library(dplyr)
#
# 1. Get the true count and percentage of all category occurrences (No distinct)
cat_counts <- PITS_deduplicated %>%
  filter(!is.na(parsed_categories1)) %>%
  count(parsed_categories1, name = "total_category_instances") %>%
  mutate(pct_of_all_categories = total_category_instances / sum(total_category_instances) * 100)
#
# 2. Get the unique incident counts and suspension rates (With distinct)
cat_suspensions <- PITS_deduplicated %>%
  filter(!is.na(parsed_categories1)) %>%
  distinct(`Incident ID`, parsed_categories1, .keep_all = TRUE) %>%
  group_by(parsed_categories1) %>%
  summarise(
    unique_incidents = n(),
    n_suspended = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE) * 100, # converted to %
    .groups = "drop"
  )

# 3. Combine them into the final table
categories_table <- cat_counts %>%
  left_join(cat_suspensions, by = "parsed_categories1") %>%
  arrange(desc(unique_incidents))

print(categories_table)

# Proportion of incidents by infraction category
category_counts <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>%
  filter(!is.na(parsed_categories1)) %>%
  count(parsed_categories1) %>%
  mutate(pct = n / sum(n) * 100) %>%
  arrange(desc(n))

# # Suspension rate *within* each category
# # (are large categories also suspending at higher rates?)
# PITS_deduplicated %>%
#   distinct(`Incident ID`, .keep_all = TRUE) %>%
#   filter(!is.na(parsed_categories1)) %>%
#   group_by(parsed_categories1) %>%
#   summarise(
#     n               = n(),
#     n_suspended     = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   ) %>%
#   arrange(desc(n))
# 

```

```{r}
# table(PITS_deduplicated$parsed_categories1)
```

Tests whether suspension length varies by infraction category — if it doesn't, lengths aren't being calibrated to severity either
### too logical inane

```{r}
# # Filter to suspended incidents only, with valid length
susp_length_df <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>%
  filter(susp_binary == 1,
         !is.na(Suspension_length),
         !is.na(parsed_categories1))

# Descriptive: median length by infraction category
length_by_infraction <- susp_length_df %>%
  group_by(parsed_categories1) %>%
  summarise(
    n              = n(),
    median_days    = median(Suspension_length),
    mean_days      = mean(Suspension_length),
    sd_days        = sd(Suspension_length),
    .groups = "drop"
  ) %>%
  arrange(desc(median_days))

print(length_by_infraction)

# Kruskal-Wallis test
kw_length <- kruskal.test(
  Suspension_length ~ as.factor(parsed_categories1),
  data = susp_length_df
)
print(kw_length)

# Dunn post-hoc test (only meaningful if KW is significant,
# but run regardless — non-significant pairs support your argument)
dunn_length <- dunn.test(
  x     = susp_length_df$Suspension_length,
  g     = susp_length_df$parsed_categories1,
  method = "bonferroni",
  kw    = TRUE
)

# Kruskal-Wallis by location
kw_location <- kruskal.test(
  Suspension_length ~ as.factor(Location_masked),
  data = susp_length_df
)
print(kw_location)
```

Define severity of categories
```{r}
# Define severity from highest to lowest
# severity_hierarchy <- c(
#   "Dangerous, Threatening, or Illegal Conduct", 
#   "Disruptive Personal Behavior", 
#   "Use and preservation of library materials and property", 
#   "Incident Impact Level - Staff Traumatic Stress", 
#   "Other", 
#   "Children in the Library", 
#   "No Infractions Set", 
#   "Medical 911"
# )
```

Code modified from Google AI
```{r}
library(dplyr)
library(dunn.test)

# 1. Isolate one dominant category per Incident ID based on severity
susp_length_dominant <- PITS_deduplicated %>% 
  filter(
    susp_binary == 1, 
    !is.na(Suspension_length), 
    !is.na(parsed_categories1),
    # parsed_categories %in% severity_hierarchy # ensures category matches your list
  ) %>% 
  # Convert to a factor matching your hierarchy order
  # mutate(parsed_categories = factor(parsed_categories, levels = severity_hierarchy, ordered = TRUE)) %>% 
  # Arrange by ID, then by the highest severity tier (lowest factor numeric value)
  arrange(`Incident ID`, parsed_categories1) %>% 
  # Keep only the first (most severe) row for each unique Incident ID
  distinct(`Incident ID`, .keep_all = TRUE)

# # 2. Descriptive summary table for dominant infractions
# dominant_summary <- susp_length_dominant %>% 
#   group_by(parsed_categories1) %>% 
#   summarise(
#     n = n(),
#     median_days = median(Suspension_length),
#     mean_days = mean(Suspension_length),
#     sd_days = sd(Suspension_length),
#     .groups = "drop"
#   ) %>% 
#   arrange(parsed_categories1) # lists from highest to lowest severity
# 
# print(dominant_summary)


```

```{r}
# 3. Kruskal-Wallis test (Valid: one unique incident per row)
kw_dominant <- kruskal.test(
  Suspension_length ~ parsed_categories1, 
  data = susp_length_dominant
)

print(kw_dominant)

# 4. Dunn post-hoc test
dunn_dominant <- dunn.test(
  x = susp_length_dominant$Suspension_length,
  g = susp_length_dominant$parsed_categories1,
  method = "bonferroni",
  kw = TRUE
)

print(dunn_dominant)
```
rework with location masked # not so significant with the p values in the DUNN
```{r}
# 1. Isolate one dominant category per Incident ID based on severity
susp_length_dominant <- PITS_deduplicated %>% 
  filter(
    susp_binary == 1, 
    !is.na(Suspension_length), 
    !is.na(Location_masked),
    # parsed_categories %in% severity_hierarchy # ensures category matches your list
  ) %>% 
  # Convert to a factor matching your hierarchy order
  # mutate(parsed_categories = factor(parsed_categories, levels = severity_hierarchy, ordered = TRUE)) %>% 
  # Arrange by ID, then by the highest severity tier (lowest factor numeric value)
  arrange(`Incident ID`, Location_masked) %>% 
  # Keep only the first (most severe) row for each unique Incident ID
  distinct(`Incident ID`, .keep_all = TRUE)

# 2. Descriptive summary table for dominant infractions
dominant_summary <- susp_length_dominant %>% 
  group_by(Location_masked) %>% 
  summarise(
    n = n(),
    median_days = median(Suspension_length),
    mean_days = mean(Suspension_length),
    sd_days = sd(Suspension_length),
    .groups = "drop"
  ) %>% 
  arrange(Location_masked) # lists from highest to lowest severity

print(dominant_summary)

# 3. Kruskal-Wallis test (Valid: one unique incident per row)
kw_dominant <- kruskal.test(
  Suspension_length ~ Location_masked, 
  data = susp_length_dominant
)

print(kw_dominant)

# 4. Dunn post-hoc test
dunn_dominant <- dunn.test(
  x = susp_length_dominant$Suspension_length,
  g = susp_length_dominant$Location_masked,
  method = "bonferroni",
  kw = TRUE
)

print(dunn_dominant)
```


DO NOT USE
# =============================================================================
# SECTION 5 Poisson/NB regression — usage volume does not
#                          predict incident counts (support)
# =============================================================================
# =============================================================================
# Pre-compute ratio data (computer users vs incidents)
# =============================================================================
DO NOT USE

```{r}
# PITS_incident_counts1 <- PITS_deduplicated %>%
#   # distinct(`Incident ID`, .keep_all = TRUE) %>% 
#   group_by(Location_masked, date = as.Date(Date2)) %>%
#   summarise(incident_count = n_distinct(`Incident ID`), .groups = "drop")
# 
# MyPC_counts <- MyPC_public %>%
#   distinct(`Login ID`, .keep_all = TRUE) %>% 
#   group_by(Location_masked, date = booking_date) %>%
#   summarise(user_count = n(), .groups = "drop")
# 
# ratio_data1 <- MyPC_counts %>%
#   full_join(PITS_incident_counts1, by = c("Location_masked", "date")) 
# # %>%
#   mutate(
#     user_count     = replace_na(user_count,     0),
#     incident_count = replace_na(incident_count, 0)
#   )

```


```{r}
#| warning: false

# # Poisson model: do more computer users → more incidents?
# poisson_model <- glm(
#   incident_count ~ user_count + Location_masked,
#   data   = ratio_data1,
#   family = poisson
# )
# summary(poisson_model) # ratio data only has one row per incident but no accurate data on categories
# 
# # Check for overdispersion (variance >> mean → use negative binomial)
# dispersion_ratio <- sum(residuals(poisson_model, type = "pearson")^2) /
#                     poisson_model$df.residual
# cat("Dispersion ratio:", round(dispersion_ratio, 3),
#     " (>1.5 suggests overdispersion → prefer NB)\n")
# 
# # Negative binomial model (more robust if overdispersed)
# nb_model <- glm.nb(
#   incident_count ~ user_count + Location_masked,
#   data = ratio_data
# )
# summary(nb_model)
# 
# # Tidy both for comparison
# tidy_poisson <- tidy(poisson_model, exponentiate = TRUE, conf.int = TRUE)
# tidy_nb      <- tidy(nb_model,      exponentiate = TRUE, conf.int = TRUE)
# 
# cat("\n--- Poisson IRRs ---\n"); print(tidy_poisson)
# cat("\n--- Negative Binomial IRRs ---\n"); print(tidy_nb)
```


# =============================================================================
# SECTION 6: Chi-square — suspension rates vary by location (inconsistent
#            application)
# =============================================================================

```{r}
#| warning: false
#| message: false

# PITS_public_ID <- PITS_public %>% 
#   distinct(`Incident ID`, .keep_all = TRUE)

# Contingency table: location × suspension binary
# location_susp_table <- table(
#   location  = PITS_deduplicated$Location_masked,
#   suspended = PITS_deduplicated$susp_binary
# )
# 
# print(location_susp_table)

# chi_location <- chisq.test(location_susp_table)
# print(chi_location)

```
### Chi-sq is significnt for location

## repeated logit regression for location
```{r}
# 1. Deduplicate to one row per Incident ID (all incidents, not just suspended ones)
incident_level_df <- PITS_deduplicated %>%
  filter(!is.na(susp_binary), !is.na(Location_masked)) %>%
  distinct(`Incident ID`, .keep_all = TRUE)

# 2. Contingency table: location × suspension binary
location_susp_table <- table(
  location  = incident_level_df$Location_masked,
  suspended = incident_level_df$susp_binary
)
print(location_susp_table)

# 3. Chi-square test — is suspension likelihood associated with location?
chi_location <- chisq.test(location_susp_table)
print(chi_location)

# 4. Cramér's V — effect size
library(rcompanion)
cramer_location <- cramerV(location_susp_table)
cat("Cramér's V — Location:", cramer_location, "\n")

# 5. Logistic regression — location as predictor of suspension (binary outcome)
logit_location <- glm(
  susp_binary ~ Location_masked,
  data   = incident_level_df,
  family = binomial
)
summary(logit_location)
```


## logit for hour
```{r}
# 1. Deduplicate to one row per Incident ID (all incidents, not just suspended ones)
incident_level_df <- PITS_deduplicated %>%
  filter(!is.na(susp_binary), !is.na(hour_label)) %>%
  distinct(`Incident ID`, .keep_all = TRUE)

# 2. Contingency table: location × suspension binary
hour_table <- table(
  location  = incident_level_df$hour_label,
  suspended = incident_level_df$susp_binary
)
print(location_hour_table)

# 3. Chi-square test — is suspension likelihood associated with location?
chi_hour <- chisq.test(hour_table)
print(chi_hour)

# 4. Cramér's V — effect size
library(rcompanion)
cramer_hour <- cramerV(hour_table)
cat("Cramér's V — Location:", cramer_hour, "\n")

# 5. Logistic regression — location as predictor of suspension (binary outcome)
logit_hour <- glm(
  susp_binary ~ hour_label,
  data   = incident_level_df,
  family = binomial
)
summary(logit_hour)
```
### and the chi-sq for the hour-label is signifnicant

```{r}
# Standardised residuals — which locations suspend more/less than expected?
std_residuals <- as.data.frame(chi_location$stdres) %>%
  setNames(c("location", "suspended", "std_resid")) %>%
  filter(suspended == "1") %>%
  arrange(desc(std_resid))

cat("\nStandardised residuals (suspended == 1):\n")
print(std_residuals)

# Logistic regression: location as predictor of suspension
logit_location <- glm(
  susp_binary ~ Location_masked,
  data   = PITS_deduplicated,
  family = binomial
)
summary(logit_location)
tidy(logit_location, exponentiate = TRUE, conf.int = TRUE)

```


# =============================================================================
# SECTION 7: Day of week and time of day — incidents and suspensions
# =============================================================================

```{r}
#| warning: false
#| message: false

# ── 7a. Order weekdays Mon–Sun ────────────────────────────────────────────────
# weekday_levels <- c("Monday", "Tuesday", "Wednesday", "Thursday",
#                     "Friday", "Saturday", "Sunday")
# 
# PITS_public <- PITS_public %>%
#   mutate(PITS_weekday = factor(PITS_weekday, levels = weekday_levels))

# ── 7b. Incidents and suspensions by day of week ──────────────────────────────
dow_summary <- PITS_deduplicated %>%
  filter(!is.na(PITS_weekday)) %>%
  group_by(PITS_weekday) %>%
  summarise(
    n_incidents   = n_distinct(`Incident ID`),
    n_suspensions = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

print(dow_summary)

# Chi-square: are incidents distributed evenly across days?
chisq_dow_incidents <- chisq.test(dow_summary$n_incidents)
cat("\nChi-square — incident counts across weekdays:\n")
print(chisq_dow_incidents)

# Chi-square: are suspensions distributed evenly across days?
chisq_dow_suspensions <- chisq.test(dow_summary$n_suspensions)
cat("\nChi-square — suspension counts across weekdays:\n")
print(chisq_dow_suspensions)

# ── 7c. Suspension rate by day — is any day more likely to suspend? ───────────
# Logistic regression with weekday as predictor
logit_dow <- glm(
  susp_binary ~ PITS_weekday,
  data   = PITS_deduplicated %>% filter(!is.na(PITS_weekday)),
  family = binomial
)
cat("\nLogistic regression — weekday as predictor of suspension:\n")
tidy(logit_dow, exponentiate = TRUE, conf.int = TRUE) %>% print()

# In Section 7b — filter Sunday out
dow_summary <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  # filter(!is.na(PITS_weekday), PITS_weekday != "Sunday") %>%
  group_by(PITS_weekday) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

# In Section 7c — filter Sunday out of the logistic regression too
logit_dow <- glm(
  susp_binary ~ PITS_weekday,
  data   = PITS_deduplicated %>% filter(!is.na(PITS_weekday), PITS_weekday != "Sunday"),
  family = binomial
)
# ── 7d. Incidents and suspensions by hour of day ──────────────────────────────
hour_summary <- PITS_deduplicated %>%
  filter(!is.na(hour_label)) %>%
  group_by(hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(hour_label)

print(dow_summary)

```


```{r}
library(lubridate)


hour_summary <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(hour_label)) %>%
  mutate(hour_label = substr(hour_label, 1, 2)) %>%
  group_by(hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(as.numeric(hour_label))


# Kruskal-Wallis: does suspension rate vary by hour?
# (use raw binary per row, not aggregated rates)
kw_hour <- kruskal.test(
  susp_binary ~ as.factor(hour_label),
  data = PITS_deduplicated %>% filter(!is.na(hour_label))
)
cat("\nKruskal-Wallis — suspension rate by hour of day:\n")
print(kw_hour)

# ── 7e. Cross-tab: day of week × hour of day (incident counts) ───────────────
dow_hour_table <- PITS_deduplicated %>%
  distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(PITS_weekday), !is.na(hour_label)) %>%
  group_by(PITS_weekday, hour_label) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

# Peak combinations — top 10 by incident count
# cat("\nTop 10 day × hour combinations by incident count:\n")
dow_hour_table <- dow_hour_table %>%
  arrange(desc(n_incidents)) %>%
  slice_head(n = 10) %>%
  print()

# Peak combinations — top 10 by suspension rate (min 5 incidents to filter noise)
cat("\nTop 10 day × hour combinations by suspension rate (n ≥ 5):\n")
dow_hour_table <- dow_hour_table %>%
  filter(n_incidents >= 5) %>%
  arrange(desc(suspension_rate)) %>%
  slice_head(n = 10) %>%
  print()
```



```{r}
incident_hour_table <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(PITS_weekday), !is.na(hour_label)) %>%
  group_by(PITS_weekday, hour_label) %>%
  summarise(
    n_incidents = n_distinct(`Incident ID`),
    .groups = "drop"
  ) %>%
  mutate(
    incident_rate = n_incidents / sum(n_incidents)
  ) 

incident_hour_table%>%
  arrange(desc(incident_rate)) %>%
  slice_head(n = 10)
```

```{r}
incident_hour_table
```


```{r}
# ── 8a. Summarise MyPC usage by day of week ───────────────────────────────────
# MyPC_public <- MyPC_public %>%
#   mutate(MyPC_weekday = factor(MyPC_weekday, levels = weekday_levels))
# 
# mypc_dow <- MyPC_public %>%
#   filter(!is.na(MyPC_weekday)) %>%
#   group_by(MyPC_weekday) %>%
#   summarise(
#     n_sessions    = n(),
#     n_unique_users = n_distinct(new_login_id),
#     .groups = "drop"
#   )

# filter out sunday
mypc_dow <- MyPC_public %>%
  distinct(new_login_id, .keep_all = TRUE) %>%
  filter(!is.na(MyPC_weekday), MyPC_weekday != "Sunday") %>%
  group_by(MyPC_weekday) %>%
  summarise(
    n_sessions     = n(),
    n_unique_users = n_distinct(new_login_id),
    .groups = "drop"
  )
print(mypc_dow)



```


```{r}
# ── 8b. Join MyPC usage with incident counts by weekday ───────────────────────
# dow_combined <- dow_summary %>%
dow_combined <- dow_summary %>%
  left_join(mypc_dow, by = c("PITS_weekday" = "MyPC_weekday")) %>%
  mutate(
    incidents_per_100_sessions = (n_incidents / n_sessions) * 100,
    suspensions_per_100_sessions = (n_suspensions / n_sessions) * 100
  )

print(dow_combined)
```


```{r}
# ── 8c. Correlation: do busier days produce more incidents? ───────────────────
cor_sessions_incidents <- cor.test(
  dow_combined$n_sessions,
  dow_combined$n_incidents,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. incidents by weekday:\n")
print(cor_sessions_incidents)

cor_sessions_suspensions <- cor.test(
  dow_combined$n_sessions,
  dow_combined$n_suspensions,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. suspensions by weekday:\n")
print(cor_sessions_suspensions)

```




```{r fig.width = 9, fig.height = 4}
#| warning: false
#| message: false
# ── 8d. Correlation: does higher usage predict higher suspension RATE? ─────────
# the p value is too high to say that rate of pc use increase affects suspension
cor_sessions_rate <- cor.test(
  dow_combined$n_sessions,
  dow_combined$suspension_rate,
  method = "spearman"
)
cat("\nSpearman correlation — sessions vs. suspension rate by weekday:\n")
print(cor_sessions_rate)
```


Check duplicate ids

```{r}
library(dplyr)
# Tidyverse approach to isolate all duplicates
all_dupes <- PITS_public %>%
  filter(`Incident ID` %in% `Incident ID`[duplicated(`Incident ID`)]) %>%
  arrange(`Incident ID`)

# # Keep only the ID and columns that actually contain differences
# varying_data <- all_dupes %>%
#   group_by(`Incident ID`) %>%
#   filter(any(n_distinct(across(everything())) > 1)) %>% # checks if rows differ at all
#   select(`Incident ID`, where(~ any(n_distinct(.) > 1)))
# 
# print(varying_data)
```



moved up to an earlier section
```{r}
# get seasons
# PITS_public <- PITS_public %>%
#   mutate(
#     month_num   = as.integer(format(Date2, "%m")),
#     month_name  = factor(format(Date2, "%B"),
#                          levels = month.name),  # orders Jan–Dec correctly
#     season      = case_when(
#       month_num %in% c(6, 7, 8)  ~ "Summer",
#       month_num %in% c(9, 10, 11) ~ "Fall",
#       month_num %in% c(12, 1, 2)  ~ "Winter",
#       month_num %in% c(3, 4, 5)   ~ "Spring",
#       TRUE ~ NA_character_
#     ),
#     season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
#   )
# 
# 
# month_summary <- PITS_public %>%
#   filter(!is.na(month_name)) %>%
#   group_by(month_name) %>%
#   summarise(
#     n_incidents     = n_distinct(`Incident ID`),
#     n_suspensions   = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# 
# season_summary <- PITS_public %>%
#   filter(!is.na(season)) %>%
#   group_by(season) %>%
#   summarise(
#     n_incidents     = n_distinct(`Incident ID`),
#     n_suspensions   = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# print(season_summary)
```

1. Add cramer's v test 
```{r}
library(rcompanion)
# PITS_public2 <- PITS_public %>% 
#    distinct(`Incident ID`, .keep_all = TRUE)

location_susp_table <- table(
  location  = PITS_deduplicated$Location_masked,
  suspended = PITS_deduplicated$susp_binary
)


# print(location_susp_table)
# Location × suspension
cramer_location <- cramerV(location_susp_table)
cat("Cramér's V — Location:", cramer_location, "\n")

# season × suspension
season_susp_table <- table(
  season   = PITS_deduplicated$season,
  suspended = PITS_deduplicated$susp_binary
)
cramer_season <- cramerV(season_susp_table)
cat("Cramér's V — Season:", cramer_season, "\n")

# month × suspension
month_susp_table <- table(
  month   = PITS_deduplicated$month_name,
  suspended = PITS_deduplicated$susp_binary
)
cramer_month <- cramerV(month_susp_table)
cat("Cramér's V — Month:", cramer_month, "\n")

# Weekday × suspension
weekday_susp_table <- table(
  weekday   = PITS_deduplicated$PITS_weekday,
  suspended = PITS_deduplicated$susp_binary
)
cramer_weekday <- cramerV(weekday_susp_table)
cat("Cramér's V — Weekday:", cramer_weekday, "\n")

# Hour × suspension
hour_susp_table <- table(
  hour      = PITS_deduplicated$hour_label,
  suspended = PITS_deduplicated$susp_binary
)
cramer_hour <- cramerV(hour_susp_table)
cat("Cramér's V — Hour:", cramer_hour, "\n")
```


2. Combined logistic regression — compare contribution of each predictor
This is the key test. Put location, weekday, and hour into the same model and see which one contributes the most explanatory power when the others are controlled for.
```{r}
combined_df <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label))

# Full model
logit_combined <- glm(
  susp_binary ~ Location_masked + PITS_weekday + hour_label,
  data   = combined_df,
  family = binomial
)
summary(logit_combined)

# Drop each block and compare fit via likelihood ratio test
logit_no_location <- glm(susp_binary ~ PITS_weekday + hour_label,
                         data = combined_df, family = binomial)
logit_no_weekday   <- glm(susp_binary ~ Location_masked + hour_label,
                         data = combined_df, family = binomial)
logit_no_hour       <- glm(susp_binary ~ Location_masked + PITS_weekday,
                         data = combined_df, family = binomial)

cat("\nDoes removing Location hurt the model? (larger stat = location matters more)\n")
anova(logit_no_location, logit_combined, test = "LRT")

cat("\nDoes removing Weekday hurt the model?\n")
anova(logit_no_weekday, logit_combined, test = "LRT")

cat("\nDoes removing Hour hurt the model?\n")
anova(logit_no_hour, logit_combined, test = "LRT")
```

collapse closed hours
```{r}
# combined_df %>%
#   count(Location_masked, susp_binary) %>%
#   pivot_wider(names_from = susp_binary, values_from = n, values_fill = 0) %>%
#   filter(`1` < 3 | `0` < 3)  # flag branches with near-zero cells
# 
# combined_df %>%
#   count(hour_label, susp_binary) %>%
#   pivot_wider(names_from = susp_binary, values_from = n, values_fill = 0) %>%
#   filter(`1` < 3 | `0` < 3)  # flag hours with near-zero cells
```
## Exclude hours and Sunday
```{r}
# Hours to exclude: 21:00–23:00 and 00:00–09:00 (keep only 9:00–20:00)
excluded_hours <- sprintf("%02d:00:00", c(21:23, 0:9))

# levels(PITS_deduplicated$hour_label)
# excluded_hours %in% levels(PITS_deduplicated$hour_label)

combined_df_exc <- PITS_deduplicated %>%
  filter(PITS_weekday != "Sunday") %>% 
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label),
         !hour_label %in% excluded_hours) %>%
  mutate(hour_label = droplevels(hour_label))  # drop the now-unused factor levels

# Full model
logit_combined_exc <- glm(
  susp_binary ~ Location_masked + PITS_weekday + hour_label,
  data   = combined_df_exc,
  family = binomial
)
summary(logit_combined_exc)
```

Rerun with EXCLUDED hours and no Sunday
```{r}

# Full model
# logit_combined <- glm(
#   susp_binary ~ Location_masked + PITS_weekday + hour_label,
#   data   = combined_df,
#   family = binomial
# )
# summary(logit_combined)

# Drop each block and compare fit via likelihood ratio test
logit_no_location_exc <- glm(susp_binary ~ PITS_weekday + hour_label,
                         data = combined_df_exc, family = binomial)
logit_no_weekday_exc   <- glm(susp_binary ~ Location_masked + hour_label,
                         data = combined_df_exc, family = binomial)
logit_no_hour_exc       <- glm(susp_binary ~ Location_masked + PITS_weekday,
                         data = combined_df_exc, family = binomial)

cat("\nDoes removing Location hurt the model (excluded)? (larger stat = location matters more)\n")
anova(logit_no_location_exc, logit_combined_exc, test = "LRT")

cat("\nDoes removing Weekday hurt the model (excluded)?\n")

anova(logit_no_weekday_exc, logit_combined_exc, test = "LRT")

cat("\nDoes removing Hour hurt the model (excluded)?\n")
anova(logit_no_hour_exc, logit_combined_exc, test = "LRT")
```


Next: Cramér's V and pseudo-R² for the final piece
Once the separation issue is addressed, re-run:
```{r}
mcfadden <- function(model, null) {
  1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
}

logit_null     <- glm(susp_binary ~ 1, data = combined_df, family = binomial)
logit_location <- glm(susp_binary ~ Location_masked, data = combined_df, family = binomial)
logit_weekday  <- glm(susp_binary ~ PITS_weekday,     data = combined_df, family = binomial)
logit_hour     <- glm(susp_binary ~ hour_label,        data = combined_df, family = binomial)

cat("Pseudo-R² — Location alone:", round(mcfadden(logit_location, logit_null), 4), "\n")
cat("Pseudo-R² — Weekday alone: ", round(mcfadden(logit_weekday, logit_null), 4), "\n")
cat("Pseudo-R² — Hour alone:    ", round(mcfadden(logit_hour, logit_null), 4), "\n")
```

3. compare variance explained with pseudo R^2 by predictor alone
```{r}
logit_null      <- glm(susp_binary ~ 1, data = combined_df, family = binomial)
logit_location  <- glm(susp_binary ~ Location_masked, data = combined_df, family = binomial)
logit_weekday   <- glm(susp_binary ~ PITS_weekday,     data = combined_df, family = binomial)
logit_hour      <- glm(susp_binary ~ hour_label,        data = combined_df, family = binomial)

mcfadden <- function(model, null) {
  1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
}

cat("Pseudo-R² — Location alone:", round(mcfadden(logit_location, logit_null), 4), "\n")
cat("Pseudo-R² — Weekday alone: ", round(mcfadden(logit_weekday, logit_null), 4), "\n")
cat("Pseudo-R² — Hour alone:    ", round(mcfadden(logit_hour, logit_null), 4), "\n")
```



# =============================================================================
# SECTION 9: Month / Season — do summer months lead to more suspensions?
# =============================================================================

```{r}
# ── 9a. Extract month from Date2 ──────────────────────────────────────────────
# PITS_public_ID <- PITS_public_ID %>%
#   # distinct(`Incident ID`, .keep_all = TRUE) %>% 
#   mutate(
#     month_num   = as.integer(format(Date2, "%m")),
#     month_name  = factor(format(Date2, "%B"),
#                          levels = month.name),  # orders Jan–Dec correctly
#     season      = case_when(
#       month_num %in% c(6, 7, 8)  ~ "Summer",
#       month_num %in% c(9, 10, 11) ~ "Fall",
#       month_num %in% c(12, 1, 2)  ~ "Winter",
#       month_num %in% c(3, 4, 5)   ~ "Spring",
#       TRUE ~ NA_character_
#     ),
#     season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
#   )

# ── 9b. Incidents and suspensions by month ────────────────────────────────────
month_summary <- PITS_deduplicated %>%
  # distinct(`Incident ID`, .keep_all = TRUE) %>% 
  filter(!is.na(month_name)) %>%
  group_by(month_name) %>%
  summarise(
    n_incidents     = n_distinct(`Incident ID`),
    n_suspensions   = sum(susp_binary, na.rm = TRUE),
    suspension_rate = mean(susp_binary, na.rm = TRUE),
    .groups = "drop"
  )

print(month_summary)


```

```{r}
# ── 9c. Incidents and suspensions by season ───────────────────────────────────
# season_summary <- PITS_public_ID %>%
#   # distinct(`Incident ID`, .keep_all = TRUE) %>% 
#   filter(!is.na(season)) %>%
#   group_by(season) %>%
#   summarise(
#     n_incidents     = n_distinct(`Incident ID`),
#     n_suspensions   = sum(susp_binary, na.rm = TRUE),
#     suspension_rate = mean(susp_binary, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# print(season_summary)

# ── 9d. Chi-square — are incidents/suspensions evenly distributed by month? ───
# month_susp_table <- table(
#   month     = PITS_deduplicated$month_name,
#   suspended = PITS_deduplicated$susp_binary
# )
# 
# chi_month <- chisq.test(month_susp_table)
# cat("\nChi-square — month × suspension:\n")
# print(chi_month)
# 
# # Cramér's V for comparison against location/hour/weekday
# library(rcompanion)
# cramer_month <- cramerV(month_susp_table)
# cat("Cramér's V — Month:", cramer_month, "\n")

```



```{r}
# ── 9e. Season as a simpler two-level test (Summer vs. rest) ─────────────────
# PITS_public_ID <- PITS_public_ID %>%
#   # distinct(`Incident ID`, .keep_all = TRUE) %>% 
#   mutate(is_summer = if_else(season == "Summer", "Summer", "Non-Summer"))
# 
# summer_table <- table(
#   is_summer = PITS_public_ID$is_summer,
#   suspended = PITS_public_ID$susp_binary
# )
# print(summer_table)
# 
# chi_summer <- chisq.test(summer_table)
# cat("\nChi-square — Summer vs. Non-Summer:\n")
# print(chi_summer)
# 
# # Proportion test — is summer suspension rate significantly different?
# summer_rates <- PITS_public_ID %>%
#   filter(!is.na(is_summer)) %>%
#   group_by(is_summer) %>%
#   summarise(n = n(), n_suspended = sum(susp_binary, na.rm = TRUE), .groups = "drop")
# 
# prop_test_summer <- prop.test(
#   x = summer_rates$n_suspended,
#   n = summer_rates$n
# )
# cat("\nProportion test — Summer vs. Non-Summer suspension rate:\n")
# print(prop_test_summer)

```



```{r}
# ── 9f. Logistic regression — month as predictor ──────────────────────────────
logit_month <- glm(
  susp_binary ~ month_name,
  data   = PITS_public_ID %>% filter(!is.na(month_name)),
  family = binomial
)
cat("\nLogistic regression — month as predictor of suspension:\n")
tidy(logit_month, exponentiate = TRUE, conf.int = TRUE) %>% print()

# ── 9g. Does month matter once location/weekday/hour are controlled for? ─────
combined_df_month <- PITS_public_ID %>%
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label), !is.na(month_name))

logit_no_month <- glm(
  susp_binary ~ Location_masked + PITS_weekday + hour_label,
  data = combined_df_month, family = binomial
)
logit_with_month <- glm(
  susp_binary ~ Location_masked + PITS_weekday + hour_label + month_name,
  data = combined_df_month, family = binomial
)

cat("\nDoes adding Month improve the model (controlling for location/weekday/hour)?\n")
anova(logit_no_month, logit_with_month, test = "LRT")

# Pseudo-R² for month alone, for direct comparison to your other predictors
logit_null_month <- glm(susp_binary ~ 1, data = combined_df_month, family = binomial)
logit_month_alone <- glm(susp_binary ~ month_name, data = combined_df_month, family = binomial)
mcfadden <- function(model, null) 1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
cat("Pseudo-R² — Month alone:", round(mcfadden(logit_month_alone, logit_null_month), 4), "\n")
```




# Predictor Cramers_V Pseudo_R2_Nagelkerke

```{r}
library(dplyr)
library(purrr)
library(rstatix)
library(performance)

# 1. Define your outcome and your list of predictor variables
outcome_var <- "susp_binary"
predictors  <- c("parsed_categories1", "Location_masked", 
                 "hour_label", "PITS_weekday")

# 2. Map through predictors to build the metrics table
metrics_table <- map_df(predictors, function(pred) {
  
  # Clean data subset removing NAs for the specific pair to avoid bias
  clean_df <- PITS_deduplicated %>% 
    filter(!is.na(.data[[outcome_var]]), !is.na(.data[[pred]]))
  
  # Calculate Cramér's V (bivariate nominal association)
  # Convert numeric predictors to factors for the contingency table
  v_val <- clean_df %>% 
    select(all_of(c(outcome_var, pred))) %>% 
    mutate(across(everything(), as.factor)) %>% 
    table() %>% 
    rstatix::cramer_v()
  
  # Calculate Pseudo R-squared (via a single-variable GLM model)
  model_formula <- as.formula(paste(outcome_var, "~", pred))
  fit <- glm(model_formula, data = clean_df, family = binomial(link = "logit"))
  
  pseudo_r2 <- as.numeric(performance::r2_nagelkerke(fit))
  
  # Return data frame row
  data.frame(
    Predictor = pred,
    Cramers_V = round(v_val, 4),
    Pseudo_R2_Nagelkerke = round(pseudo_r2, 4)
  )
})

# 3. View table ranked by predictive strength
metrics_table <- metrics_table %>% arrange(desc(Pseudo_R2_Nagelkerke))
print(metrics_table)
```

# =============================================================================
# SECTION 10: Mixed-effects model with season added
# =============================================================================

```{r}
library(lme4)
library(performance)

# Define hours to exclude (midnight through 7am)
excluded_hours <- sprintf("%02d:00:00", 00:7)


# Prepare data — ensure season, weekday, hour, and location are all present
mixed_df <- PITS_public_ID %>%
  filter(!is.na(susp_binary), !is.na(Location_masked),
         !is.na(PITS_weekday), !is.na(hour_label), !is.na(season)) %>% 
  filter(PITS_weekday != "Sunday",
         !hour_label %in% excluded_hours)

# ── Mixed model: weekday + hour + season as fixed effects, branch as random ──
mixed_model_season <- glmer(
  susp_binary ~ PITS_weekday + hour_label + season + (1 | Location_masked),
  data    = mixed_df,
  family  = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(mixed_model_season)

# ── ICC — how much variance is due to branch, controlling for weekday/hour/season ──
icc_season <- icc(mixed_model_season)
icc_season

# ── Compare to the original model (weekday + hour only, no season) ───────────
mixed_model_original <- glmer(
  susp_binary ~ PITS_weekday + hour_label + (1 | Location_masked),
  data   = mixed_df,   # same filtered data, for fair comparison
  family = binomial
)
icc_original <- icc(mixed_model_original)

cat("\nICC comparison:\n")
cat("Without season — Adjusted ICC:", round(icc_original$ICC_adjusted, 4),
    "| Unadjusted ICC:", round(icc_original$ICC_unadjusted, 4), "\n")
cat("With season    — Adjusted ICC:", round(icc_season$ICC_adjusted, 4),
    "| Unadjusted ICC:", round(icc_season$ICC_unadjusted, 4), "\n")

# ── Likelihood ratio test — does adding season improve the mixed model? ──────
anova_season <- anova(mixed_model_original, mixed_model_season, test = "LRT")
cat("\nDoes adding season improve the mixed model?\n")
anova_season
```



```{r}
# ── Extract and rank branch-level random effects (with season controlled) ────
ranef_branch_season <- ranef(mixed_model_season)$Location_masked
ranef_branch_season$Location_masked <- rownames(ranef_branch_season)
names(ranef_branch_season)[1] <- "random_effect"

ranef_branch_season <- ranef_branch_season %>%
  arrange(desc(random_effect))

ranef_branch_season
```

